### 2017-11-05 report/draft.tex
init

\documentclass[11pt]{article}
\usepackage[a4paper, hmargin={2.8cm, 2.8cm}, vmargin={2.5cm, 2.5cm}]{geometry}
\usepackage{eso-pic} % \AddToShipoutPicture
\usepackage{graphicx} % \includegraphics

%% Change `ku-farve` to `nat-farve` to use SCIENCE's old colors or
%% `natbio-farve` to use SCIENCE's new colors and logo.
\def \ColourPDF {include/ku-farve}

%% Change `ku-en` to `nat-en` to use the `Faculty of Science` header
\def \TitlePDF   {include/ku-en}  % University of Copenhagen

\title{
  \vspace{3cm}
  \Huge{Optimizing SasView models} \\
  \Large{Enhancing performance by rewriting kernels from OpenCL/OpenMP to Futhark}
}

\author{
  \Large{Mikkel Storgaard Knudsen}
  \\ \texttt{mikkelstorgaard@gmail.com} \\
  \\
  \Large{Supervisor}
  \\ Martin Elsman
  \\ \texttt{mael@di.ku.dk} \\
}

\date{
    \today
}

\begin{document}

\AddToShipoutPicture*{\put(0,0){\includegraphics*[viewport=0 0 700 600]{\ColourPDF}}}
\AddToShipoutPicture*{\put(0,602){\includegraphics*[viewport=0 600 700 1600]{\ColourPDF}}}

\AddToShipoutPicture*{\put(0,0){\includegraphics*{\TitlePDF}}}

\clearpage\maketitle
\thispagestyle{empty}

\newpage

%% Write your dissertation here.
\section*{Goal and motivation}
\textit{SasView} is a Small Angle Scattering (SAS) analysis package for the analysis of
1D and 2D scattering data directly in inverse space.
As \textit{SasView} is given data, it performs its analyses by running the input data
through theory models provided by the standalone module \textit{Sasmodels}.
\\
\\
The \textit{Sasmodels} models utilizes a set of kernels, which are implemented
in both OpenCL and OpenMP.
However, benchmark
comparisons\footnote{https://futhark-lang.org/performance.html} between the
parallel algorithm library \textit{Thrust} (implemented in C++/CUDA), and the
array programming language Futhark, shows that there might severe speed
improvements to gain by running these model calculations in Futhark, instead
of in OpenCL and OpenMP.
\\
\\
The goal of this project is to explore the potential speed gains obtained by
rewriting, and testing the resulting speed improvements of, a subset of the
\textit{Sasmodels} models.
The subset will be chosen so that there are at least
one ``simple'' model, one model of intermediate complexity, and one complex
model in it.
\\
\\
If rewriting the models results in a significant\footnote{I.e. a speed increase
  in order of at least a magnitude}, speedup, it will reasonable to believe that the
rest of the models in \textit{Sasmodels} can be rewritten for a similar speedup.
\end{document}

### 2017-11-06 report/draft.tex
g
a
go

@@
 
+\newcommand{\sasmodels}{\textit{SasModels}}
+\newcommand{\sasview}{\textit{SasView}}
+\newcommand{\iq}{\textit{I(q)}}
 \begin{document}
@@
 
+\section*{Abstract}
+
 \newpage
 
-%% Write your dissertation here.
-\section*{Goal and motivation}
+\section{Introduction}
+\subsection{Goal and motivation}
 \textit{SasView} is a Small Angle Scattering (SAS) analysis package for the analysis of
@@
 If rewriting the models results in a significant\footnote{I.e. a speed increase
-  in order of at least a magnitude}, speedup, it will reasonable to believe that the
-rest of the models in \textit{Sasmodels} can be rewritten for a similar speedup.
+  in order of at least a magnitude}, speedup, it will reasonable to believe that 
+  the rest of the models in \textit{Sasmodels} can be rewritten for a similar
+  speedup.
+
+\section{Demonstration of use of finished project}
+As \sasmodels is rarely used on its own, but rather as the backbone of 
+\sasview's calculation backend, I am not going to write much about 
+\sasview usage.
+Instead, I will describe how Futhark models in \sasmodels are leveraged 
+by \sasmodels' users.
+% mere
+
+\subsection{Describing \sasmodels' models}
+A \sasmodels model info file (hereafter referred to as just a model info) is a 
+Python file, that defines at least two primary properties of a 
+calculation model \textit{m}:
+\begin{itemize}
+
+  \item a function definition \texttt{I(q)}, which defines scattered intensity 
+  of the scattering vector \texttt{q}.
+
+  \item a list \texttt{parameters}, of the parameters that are used in 
+  \texttt{I(q)}. If \texttt{I(q)} does not take any parameters besides 
+  \texttt{q}, the list is left empty.
+  In the model info file, the parameters are each defined with their unit name, 
+  their default values, their value range, their type of parameter and
+  a short text describing what they are.
+\end{itemize}
+
+Additionally, the model info file also always contains the short-name, 
+the full title, a category, and a mathematical \footnote{a representation 
+that compiles to a meaningful mathematical expression if used in a LaTeX math 
+environment} string representation of the scattering function.
+
+As the scattering function \texttt{I(q)} and its parameters are the only 
+two things that are strictly necessary in a model definition, the developer
+has the option to leave the model info defined with just these six properties.
+
+\sasmodels has built-in default definitions for \texttt{I(q_x,q_y)} if working
+with two-dimensional data, and other functions such as \texttt{form_volume(q)}
+for working with voluminous scatterings.
+
+Therefore, it is optional for the developer to define a custom definition
+of \texttt{I(q_x, q_y)} or other default functions.
+An example of this can be seen in appendix dab or line.
+
+A minimum viable model is showed in figure \ref{fig:linemodel}, and a more
+complex model can be seen in appendix \ref{app:core_shell}
+
+\begin{figure}
+  % fuld beskrivelse af ax+b-figur, med parametre a, b, og en Iqxy
+  \caption{model for modelling a line}\label{fig:linemodel}
+\end{figure}
+
+\subsection{Defining \texttt{I(q)} in Python and OpenCL}
+A \sasmodels \iq function can be written in either Python (which will be
+executed in a Numpy context at runtime), or it can be written in C, which will
+will be inserted into an OpenCL template, so that it can be compiled and 
+executed as a kernel in an OpenCL context, preferably on a GPU.
+
+A Python \iq can be seen in figure \ref{fig:linemodel}, whereas both DAB (figure
+\ref{fig:dab} and core_shell(figure \ref{fig:core_shell}) is written in C.
+
+But with the arrival of Futhark models in \sasmodels, there is now a third
+additional way of writing \sasmodels models.
+
+\subsection{Using }
+
+% definition af gammel sasmodels-model
+% evt. kald af model.
+
+% til vægs
+
+\section{Ease of use for almene brugere}
+% forskel på at skrive moduler, og bare at hente et oversat modul ned
+
+\section{Implementation}
+% current architecture of Sasmodels engine
+% description of kernelcl kernels
+% (smaller) description of kernelcl kernels
+% description of current kernelfut implementation
+% designovervejelser, hvorfor selvstændige moduler, hvorfor ikke bare udvide kernelpy
+%% hvorfor ikke bare definere futhark-Iq-funktionen som string i model.py?
+% fordi der ikke er noget kodegnererering ved runtime/ingen futhark-compile-fejl
+% under runtime, og prægenererede moduler kan deles med andre, uden disse
+% ``andre''' behøver have futhark-oversætteren selv.
+
+\section{Performance tests}
+
+% description of built-in sasmodels comparison functionality
+%% description of how comparison module has dictated the design of the futhark kernel (i.e. loading as much data as possible on init)
+
+% display results
+
+% discussion of futhark vs. python performance
+%% where does the performance gains come from?
+
+% discussion of futhark vs. opencl performance
+%% where does the performance gains come from?
+%%% MASSER AF REFERENCER TIL TROELS' PAPERS
+
+% correctness
+
+\section{Discussion, future work}
+% is this immediately useful? not on single fire runs with low nq
+% can be used for data streams / bundling
+% experience gains can lay groundwork for parallelising finding best match
+
+% user directed: ease of use,
+% (evt.) skønhedsfejl
+
+\section{Conclusion}
+% nævn at projektet er blevet præsenteret for en faktisk brugergruppe,
+% og fik god og konstruktiv feedback
+% det inspirerede til yderligere projekter/forbedringer indenfor sasmodels
+
+\section*{Appendices}
+
+\subsection*{DAB model}
+
+\subsection*{broad_peak model}
+\begin{figure}
+  \begin[language=Python]{lstlisting}
+r"""
+Definition
+----------
+This model calculates an empirical functional form for SAS data characterized
+by a broad scattering peak. Many SAS spectra are characterized by a broad peak
+even though they are from amorphous soft materials. For example, soft systems
+that show a SAS peak include copolymers, polyelectrolytes, multiphase systems,
+layered structures, etc.
+The d-spacing corresponding to the broad peak is a characteristic distance
+between the scattering inhomogeneities (such as in lamellar, cylindrical, or
+spherical morphologies, or for bicontinuous structures).
+The scattering intensity $I(q)$ is calculated as
+.. math:: I(q) = \frac{A}{q^n} + \frac{C}{1 + (|q - q_0|\xi)^m} + B
+Here the peak position is related to the d-spacing as $q_0 = 2\pi / d_0$.
+$A$ is the Porod law scale factor, $n$ the Porod exponent, $C$ is the
+Lorentzian scale factor, $m$ the exponent of $q$, $\xi$ the screening length,
+and $B$ the flat background.
+For 2D data the scattering intensity is calculated in the same way as 1D,
+where the $q$ vector is defined as
+.. math:: q = \sqrt{q_x^2 + q_y^2}
+References
+----------
+None.
+Authorship and Verification
+----------------------------
+* **Author:** NIST IGOR/DANSE **Date:** pre 2010
+* **Last Modified by:** Paul kienle **Date:** July 24, 2016
+* **Last Reviewed by:** Richard Heenan **Date:** March 21, 2016
+"""
+
+from numpy import inf, errstate
+
+name = "broad_peak"
+title = "Broad Lorentzian type peak on top of a power law decay"
+description = """\
+      I(q) = scale_p/pow(q,exponent)+scale_l/
+      (1.0 + pow((fabs(q-q_peak)*length_l),exponent_l) )+ background
+      List of default parameters:
+      porod_scale = Porod term scaling
+      porod_exp = Porod exponent
+      lorentz_scale = Lorentzian term scaling
+      lorentz_length = Lorentzian screening length [A]
+      peak_pos = peak location [1/A]
+      lorentz_exp = Lorentzian exponent
+      background = Incoherent background"""
+category = "shape-independent"
+
+# pylint: disable=bad-whitespace, line-too-long
+#             ["name", "units", default, [lower, upper], "type", "description"],
+parameters = [["porod_scale",    "",  1.0e-05, [-inf, inf], "", "Power law scale factor"],
+              ["porod_exp",      "",      3.0, [-inf, inf], "", "Exponent of power law"],
+              ["lorentz_scale",  "",     10.0, [-inf, inf], "", "Scale factor for broad Lorentzian peak"],
+              ["lorentz_length", "Ang",  50.0, [-inf, inf], "", "Lorentzian screening length"],
+              ["peak_pos",       "1/Ang", 0.1, [-inf, inf], "", "Peak position in q"],
+              ["lorentz_exp",    "",      2.0, [-inf, inf], "", "Exponent of Lorentz function"],
+             ]
+# pylint: enable=bad-whitespace, line-too-long
+
+def Iq(q,
+       porod_scale=1.0e-5,
+       porod_exp=3.0,
+       lorentz_scale=10.0,
+       lorentz_length=50.0,
+       peak_pos=0.1,
+       lorentz_exp=2.0):
+    """
+    :param q:              Input q-value
+    :param porod_scale:    Power law scale factor
+    :param porod_exp:      Exponent of power law
+    :param lorentz_scale:  Scale factor for broad Lorentzian peak
+    :param lorentz_length: Lorentzian screening length
+    :param peak_pos:       Peak position in q
+    :param lorentz_exp:    Exponent of Lorentz function
+    :return:               Calculated intensity
+    """
+    z = abs(q - peak_pos) * lorentz_length
+    with errstate(divide='ignore'):
+        inten = (porod_scale / q ** porod_exp
+                 + lorentz_scale / (1 + z ** lorentz_exp))
+    return inten
+Iq.vectorized = True  # Iq accepts an array of q values
+
+def random():
+    import numpy as np
+    pars = dict(
+        scale=1,
+        porod_scale=10**np.random.uniform(-8, -5),
+        porod_exp=np.random.uniform(1, 6),
+        lorentz_scale=10**np.random.uniform(0.3, 6),
+        lorentz_length=10**np.random.uniform(0, 2),
+        peak_pos=10**np.random.uniform(-3, -1),
+        lorentz_exp=np.random.uniform(1, 4),
+    )
+    pars['lorentz_length'] /= pars['peak_pos']
+    pars['lorentz_scale'] *= pars['porod_scale'] / pars['peak_pos']**pars['porod_exp']
+    #pars['porod_scale'] = 0.
+    return pars
+
+demo = dict(scale=1, background=0,
+            porod_scale=1.0e-05, porod_exp=3,
+            lorentz_scale=10, lorentz_length=50, peak_pos=0.1, lorentz_exp=2)
+
+  \end{lstlisting}
+  \caption{The broad_peak model}
+
+\end{figure}
+
 \end{document}

### 2017-11-06 report/draft.tex
b

@@
 
-\sasmodels has built-in default definitions for \texttt{I(q_x,q_y)} if working
-with two-dimensional data, and other functions such as \texttt{form_volume(q)}
+\sasmodels has built-in default definitions for $I(q_x,q_y)$ if working
+with two-dimensional data, and other functions such as \texttt{form\_volume(q)}
 for working with voluminous scatterings.
@@
 Therefore, it is optional for the developer to define a custom definition
-of \texttt{I(q_x, q_y)} or other default functions.
+of $I(q_x, q_y)$ or other default functions.
 An example of this can be seen in appendix dab or line.
@@
   % fuld beskrivelse af ax+b-figur, med parametre a, b, og en Iqxy
-  \caption{model for modelling a line}\label{fig:linemodel}
+  \caption{model info file for modelling a line}\label{fig:linemodel}
 \end{figure}
 
-\subsection{Defining \texttt{I(q)} in Python and OpenCL}
+\subsection{Defining \iq in Python and OpenCL}
 A \sasmodels \iq function can be written in either Python (which will be
@@
 A Python \iq can be seen in figure \ref{fig:linemodel}, whereas both DAB (figure
-\ref{fig:dab} and core_shell(figure \ref{fig:core_shell}) is written in C.
+\ref{fig:dab} and core\_shell(figure \ref{fig:core-shell}) is written in C.
+Notice that the main difference between the models lies in the Iq property.
+For Python models, \iq is a callable Python function, whilst the C \iq is a
+ string which is copy-pasted wholesale into aforementioned OpenCL template.
 
-But with the arrival of Futhark models in \sasmodels, there is now a third
-additional way of writing \sasmodels models.
+But with the addition of Futhark models in \sasmodels adds a third way of
+using and defining a \sasmodels model.
 
-\subsection{Using }
+\subsection{Using a Futhark model}
+\label{sec:using-futhark}
+There is a difference between defining and using a Futhark model.
+The \sasview user who wishes to use a Futhark model in his or her work does 
+not need more than the futhark model info file, and a copy of the compiled
+Futhark kernel. In fact, the futhark model info file is identical to OpenCL- 
+and Python model info files, except for in two properties.
+As show in figure \ref{fig:futhark_dab}, there has been added a flag 
+%CHECK THISSS!!!
+\texttt{model\_is\_futhark: true}, and \iq is now a path to a precompiled 
+Futhark kernel.
+
+The user does not need to have any Futhark compiler installed if he merely wants
+to use the precompiled Futhark module: Futhark models for \sasmodels compiles to
+portable Python libraries which has numpy and pyopencl as their only 
+dependencies - and as \sasmodels has these dependencies as well, the user will
+already have these dependencies sorted out, if he has a working installation of 
+\sasview or \sasmodels.
+
+\subsection{Defining a Futhark model}
+To have a usable Futhark kernel to run in \sasmodels, it is first necessary to
+define it as a Futhark program and compile it using \texttt{futhark-pyopencl}.
+A model \textit{m} itself is defined as a parametric Futhark module\footnote{
+  A more in-depth explanation of the complete definition of a Futhark module is 
+  in \ref{sec:generating-futhark}},
+which contains an Iq-function, an Iqxy- and a form\_volume function.
+It is the model programmers responsibility to fill in all three functions.
+
+%% figure
+
+A simple example of a Futhark model is the line module, 
+as depicted in figure \ref{fig:linemodel-futhark}.
+Note that Iqxy is non-default, whilst form\_volume is the default 
+implementation.
+
+\subsubsection{On the user friendliness of Futhark models}
+As shown in figure \ref{fig:linemodel-futhark}, there are currently two small 
+nuisances in Futhark models, compared to equivalent models written in
+Python/OpenCL:
+
+With both Python and OpenCL models, the programmer can leave out Iqxy and 
+form\_volume so that \sasmodels automatically fills in the default definition
+for these two functions instead.
+
+The Futhark programmer does not have this leisure yet, and must define the two
+functions no matter what. This is admittedly a piece of annoying boiler plate,
+but it is also a thing that can be solved by compartmentalising the offending
+functions away in a parametric library module.
+
+The less removable nuisance is that the programmer must explicitly extract the
+Iq function parameters from the list \texttt{local\_values}. This must be done
+in the order they appear in the model info parameter list.
+
+Iq functions in Python and OpenCL can be defined with any number of function 
+parameters, but in Futhark, we need to adhere to the type declaration of Iq 
+(see \ref{has-iq}), which means that Iq can take only two arguments; the 
+scattering vector \texttt{q}, and the list \texttt{local\_values} of parameters
+needed for the scattering.
+
+This makes for an ugly prelude to the actual body of the Iq function, as shown
+in fig\ref{fig:linemodel-futhark}, or more critically in models like 
+core-shell \ref{fig:core-shell}.
+
+\subsection{Generating Futhark models using parametric modules and a header}
+\label{sec:generating-futhark}
+\begin{lstlisting}
+module type has_iq = {
+Iq: real -> real[] -> real // the model's scattering function.
+Iqxy: real -> real -> real[] -> real // the model's 2D scattering function
+form_volume: real[] -> real // the model's form volume function.
+}  
+\end{lstlisting}
 
@@
 % til vægs
+
+\subsubsection{Why not Futhark models as model info strings instead standalone modules?}
+% designovervejelser, hvorfor selvstændige moduler, hvorfor ikke bare udvide kernelpy
+%% hvorfor ikke bare definere futhark-Iq-funktionen som string i model.py?
+% fordi der ikke er noget kodegnererering ved runtime/ingen futhark-compile-fejl
+% under runtime, og prægenererede moduler kan deles med andre, uden disse
+% ``andre''' behøver have futhark-oversætteren selv.
 
@@
 % description of current kernelfut implementation
-% designovervejelser, hvorfor selvstændige moduler, hvorfor ikke bare udvide kernelpy
-%% hvorfor ikke bare definere futhark-Iq-funktionen som string i model.py?
-% fordi der ikke er noget kodegnererering ved runtime/ingen futhark-compile-fejl
-% under runtime, og prægenererede moduler kan deles med andre, uden disse
-% ``andre''' behøver have futhark-oversætteren selv.
 
@@
 %% where does the performance gains come from?
+% show core_shell-whatever; and elaborate on loop flattening into maps
 %%% MASSER AF REFERENCER TIL TROELS' PAPERS

### 2017-11-07 report/draft.tex
Pøls!

@@
 \newcommand{\iq}{\textit{I(q)}}
+\newcommand{\foriq}{\texttt{for\_iq}}
+\newcommand{\hasiq}{\texttt{has\_iq}}
 \begin{document}
@@
 \subsection{Defining a Futhark model}
+\label{sec:defining-futhark-model}
 To have a usable Futhark kernel to run in \sasmodels, it is first necessary to
@@
 \label{sec:generating-futhark}
-\begin{lstlisting}
-module type has_iq = {
-Iq: real -> real[] -> real // the model's scattering function.
-Iqxy: real -> real -> real[] -> real // the model's 2D scattering function
-form_volume: real[] -> real // the model's form volume function.
-}  
-\end{lstlisting}
+As mentioned in \ref{sec:defining-futhark-model}, the futhark model developer
+does only need to define three functions to be able to compile a working 
+\sasmodels Futhark model.
+However, those three functions are merely used by a parametric Futhark module 
+\texttt{for\_iq}, which takes care of the actual execution of the 
+Futhark kernel.
+Let's go through the actual implementation of the Futhark models.
 
-% definition af gammel sasmodels-model
-% evt. kald af model.
+\subsubsection{line\_futhark.fut (a Futhark model)}
+\label{sec:line-fut}
+\begin{figure}
+  % hele line_futhark.fut
+  \caption{The complete definition of the line model implemented in Futhark.}
+  \label{fig:line-fut}
+\end{figure}
+There a several things happening in this fig. \ref{fig:line-fut}, besides the 
+definition of the line Iq function.
 
-% til vægs
+\texttt{module line} takes either single- or double precision floats as a
+parameter, and then assigns the type alias \texttt{dtype} to the type of the
+input parameter. This means that we can use \texttt{dtype} in the type
+declarations throughout the module, and define the module for floats in general,
+instead of maybe having to define linef64- and linef32 modules.
 
-\subsubsection{Why not Futhark models as model info strings instead standalone modules?}
-% designovervejelser, hvorfor selvstændige moduler, hvorfor ikke bare udvide kernelpy
-%% hvorfor ikke bare definere futhark-Iq-funktionen som string i model.py?
-% fordi der ikke er noget kodegnererering ved runtime/ingen futhark-compile-fejl
-% under runtime, og prægenererede moduler kan deles med andre, uden disse
-% ``andre''' behøver have futhark-oversætteren selv.
+Then, two kernels (one for each of f32 and f64) are initialised by giving
+the parametric module \texttt{for\_iq} the line module and a data type 
+as parameters, returning \texttt{kernel\_float32} and \texttt{kernel\_float64}.
 
-\section{Ease of use for almene brugere}
-% forskel på at skrive moduler, og bare at hente et oversat modul ned
+Given these two modules, we can finally define four entry points into the
+finished model. These are the four entry points that can be called by \sasmodels
+whenever a \sasmodels experiment needs to execute calculations in a futhark
+kernel.
+Each of the entry points are simply passing on their call parameters to the 
+\texttt{run\_kernel} functions defined in the \texttt{for\_iq} module.
+Everything after the module line definition is boilerplate code, and could
+possibly defined in a less inelegant way in a later revision, but currently, it 
+is a necessary part of every Futhark model.
+
+Other models, such as core-shell %FIX\
+imports libraries so they are available in the Iq function.
+%% som vist i core shell, kan det også være nødvendigt at definere og importere 
+% ekstra biblioteker
+
+\subsubsection{header/for\_iq.fut (the Futhark kernel template)}
+\texttt{for\_iq.fut} contains the module type declaration for the \hasiq module,
+ and it contains the definition of the parametric \foriq module.
+
+The type declaration \hasiq ensures, that any model module that is
+used as a parameter for the \foriq module defines at least the functions
+Iq, Iqxy and form\_volume, as these functions are necessary for the definition
+of the Futhark kernel.
+
+Next, we have the actual definition of the module \foriq:
+\foriq takes two parameters (a parametric Futhark model), and a real type 
+to instantiate it with.
+It then instantiates the model with the real type, making the finished Iq module
+\texttt{Iq} available throughout the rest of the kernel.
+
+\begin{figure}
+  % hele for_iq.fut
+  % forklar module type
+  % forklar module 
+  \caption{The Futhark template that is used by the individual models.}
+  \label{fig:for-iq}
+\end{figure}
+
+\subsubsection{Why not implement Futhark models as model info strings 
+instead of standalone modules?}
+OpenCL kernels are generated by copy/pasting the desired \iq function body
+into a template, and then compiling the resulting source code.
+This happens at runtime in \sasmodels, which means that the kernels aren't
+ actually compiled until right before \sasmodels needs them.
+
+Futhark models eschews this design and opts instead to separate code generation
+and model usage completely.
+It would be simple to copy the design of the OpenCL kernels, and compile Futhark
+kernels on \sasmodels runtime, but this design was cast aside in favour of the
+design described in sec \ref{sec:generating-futhark}.
+
+There are three reasons for that:
+\begin{enumerate}
+  \item Removing code generation from runtime unequivocally decreases the time
+  it takes to run a \sasmodels calculation. Although time taken by code 
+  generation is not covered by \sasmodels' built-in benchmark modules, it should
+  be obvious that a program that merely executes a kernel computation has a 
+  shorter runtime than a program that has to compile the kernel before it
+  executes the kernel computation.
+
+  \item Compiling a kernel just in time brings the risk of encountering 
+  Futhark compilation errors at \sasmodels runtime. 
+  It is arguably a user experience nightmare to encounter compilation errors at 
+  runtime, which is why Futhark models keeps compilation separate from runtime.
+
+  \item Compiling Futhark models down to self-contained\footnote{depending only 
+  on numpy and pyopencl} Python libraries means that Futhark models are 
+  completely portable, and as such can be used by any \sasmodels or \sasview 
+  user, without them needing to have a Futhark compiler installed on their 
+  workstations.
+\end{enumerate}
 

### 2017-11-07 report/draft.tex
asd

@@
 \usepackage{graphicx} % \includegraphics
+\usepackage{listings} % \includegraphics
 
@@
 
+Then finally, it uses the Iq module to define the two functions 
+\texttt{run\_kernel}, and \texttt{run\_kernel\_2d}, which defines the actual 
+\sasmodels computation procedure with the scattering functions injected by 
+the Iq module.
+
 \begin{figure}
@@
   \item Compiling Futhark models down to self-contained\footnote{depending only 
-  on numpy and pyopencl} Python libraries means that Futhark models are 
+  on Numpy and Pyopencl} Python libraries means that Futhark models are 
   completely portable, and as such can be used by any \sasmodels or \sasview 
@@
 
-\section{Implementation}
+\section{Implementation of \texttt{kernelfut.py}}
+\sasmodels
+
 % current architecture of Sasmodels engine
@@
 \subsection*{DAB model}
+\begin{figure}
+  \lstinputlisting[language=Python]{../sasmodels/sasmodels/models/dab.py}
+  \label{fig:dab}
+  \caption{The DAB model}
+\end{figure}
 
-\subsection*{broad_peak model}
+\subsection*{broad\_peak model}
 \begin{figure}
-  \begin[language=Python]{lstlisting}
-r"""
-Definition
-----------
-This model calculates an empirical functional form for SAS data characterized
-by a broad scattering peak. Many SAS spectra are characterized by a broad peak
-even though they are from amorphous soft materials. For example, soft systems
-that show a SAS peak include copolymers, polyelectrolytes, multiphase systems,
-layered structures, etc.
-The d-spacing corresponding to the broad peak is a characteristic distance
-between the scattering inhomogeneities (such as in lamellar, cylindrical, or
-spherical morphologies, or for bicontinuous structures).
-The scattering intensity $I(q)$ is calculated as
-.. math:: I(q) = \frac{A}{q^n} + \frac{C}{1 + (|q - q_0|\xi)^m} + B
-Here the peak position is related to the d-spacing as $q_0 = 2\pi / d_0$.
-$A$ is the Porod law scale factor, $n$ the Porod exponent, $C$ is the
-Lorentzian scale factor, $m$ the exponent of $q$, $\xi$ the screening length,
-and $B$ the flat background.
-For 2D data the scattering intensity is calculated in the same way as 1D,
-where the $q$ vector is defined as
-.. math:: q = \sqrt{q_x^2 + q_y^2}
-References
-----------
-None.
-Authorship and Verification
-----------------------------
-* **Author:** NIST IGOR/DANSE **Date:** pre 2010
-* **Last Modified by:** Paul kienle **Date:** July 24, 2016
-* **Last Reviewed by:** Richard Heenan **Date:** March 21, 2016
-"""
+  \lstinputlisting[language=Python]{../sasmodels/sasmodels/models/broad_peak.py}
+  \caption{The broad_peak model}
+\end{figure}
 
-from numpy import inf, errstate
+\begin{figure}
+  \lstinputlisting[language=Python]{../sasmodels/sasmodels/models/line.py}
+  \caption{The line model}
+\end{figure}
 
-name = "broad_peak"
-title = "Broad Lorentzian type peak on top of a power law decay"
-description = """\
-      I(q) = scale_p/pow(q,exponent)+scale_l/
-      (1.0 + pow((fabs(q-q_peak)*length_l),exponent_l) )+ background
-      List of default parameters:
-      porod_scale = Porod term scaling
-      porod_exp = Porod exponent
-      lorentz_scale = Lorentzian term scaling
-      lorentz_length = Lorentzian screening length [A]
-      peak_pos = peak location [1/A]
-      lorentz_exp = Lorentzian exponent
-      background = Incoherent background"""
-category = "shape-independent"
+\begin{figure}
+  \lstinputlisting[]{../sasmodels/sasmodels/models/futharks/line_futhark.fut}
+  \label{fig:linemodel-futhark}
+  \caption{The line model written in Futhark}
+\end{figure}
 
-# pylint: disable=bad-whitespace, line-too-long
-#             ["name", "units", default, [lower, upper], "type", "description"],
-parameters = [["porod_scale",    "",  1.0e-05, [-inf, inf], "", "Power law scale factor"],
-              ["porod_exp",      "",      3.0, [-inf, inf], "", "Exponent of power law"],
-              ["lorentz_scale",  "",     10.0, [-inf, inf], "", "Scale factor for broad Lorentzian peak"],
-              ["lorentz_length", "Ang",  50.0, [-inf, inf], "", "Lorentzian screening length"],
-              ["peak_pos",       "1/Ang", 0.1, [-inf, inf], "", "Peak position in q"],
-              ["lorentz_exp",    "",      2.0, [-inf, inf], "", "Exponent of Lorentz function"],
-             ]
-# pylint: enable=bad-whitespace, line-too-long
-
-def Iq(q,
-       porod_scale=1.0e-5,
-       porod_exp=3.0,
-       lorentz_scale=10.0,
-       lorentz_length=50.0,
-       peak_pos=0.1,
-       lorentz_exp=2.0):
-    """
-    :param q:              Input q-value
-    :param porod_scale:    Power law scale factor
-    :param porod_exp:      Exponent of power law
-    :param lorentz_scale:  Scale factor for broad Lorentzian peak
-    :param lorentz_length: Lorentzian screening length
-    :param peak_pos:       Peak position in q
-    :param lorentz_exp:    Exponent of Lorentz function
-    :return:               Calculated intensity
-    """
-    z = abs(q - peak_pos) * lorentz_length
-    with errstate(divide='ignore'):
-        inten = (porod_scale / q ** porod_exp
-                 + lorentz_scale / (1 + z ** lorentz_exp))
-    return inten
-Iq.vectorized = True  # Iq accepts an array of q values
-
-def random():
-    import numpy as np
-    pars = dict(
-        scale=1,
-        porod_scale=10**np.random.uniform(-8, -5),
-        porod_exp=np.random.uniform(1, 6),
-        lorentz_scale=10**np.random.uniform(0.3, 6),
-        lorentz_length=10**np.random.uniform(0, 2),
-        peak_pos=10**np.random.uniform(-3, -1),
-        lorentz_exp=np.random.uniform(1, 4),
-    )
-    pars['lorentz_length'] /= pars['peak_pos']
-    pars['lorentz_scale'] *= pars['porod_scale'] / pars['peak_pos']**pars['porod_exp']
-    #pars['porod_scale'] = 0.
-    return pars
-
-demo = dict(scale=1, background=0,
-            porod_scale=1.0e-05, porod_exp=3,
-            lorentz_scale=10, lorentz_length=50, peak_pos=0.1, lorentz_exp=2)
-
-  \end{lstlisting}
-  \caption{The broad_peak model}
-
+\begin{figure}
+  \lstinputlisting[language=Python]{../sasmodels/sasmodels/models/line_futhark.py}
+  \caption{The line model info file designating a Futhark path}
 \end{figure}

### 2017-11-07 report/draft.tex
qqq

@@
 \usepackage{listings} % \includegraphics
+\usepackage{pgfplots}
 
@@
 
-\section{Performance tests}
+\section{Method and performance results}
+% description of built-in sasmodels comparison functionality
+\sasmodels contains a comparison module called \texttt{compare.py}, which
+lets the user run different calculations on the same dataset.
+This module has been the basis of all my performance tests, and indeed also
+the module that has let me design Futhark models in the first place.
 
-% description of built-in sasmodels comparison functionality
-%% description of how comparison module has dictated the design of the futhark kernel (i.e. loading as much data as possible on init)
+\subsection{Development method}
+The development of Futhark for \sasmodels has been largely test-driven.
+For any given already existing model in the \sasmodels catalogue that I have
+wished to implement in Futhark, I have started out by defining an equivalent
+model written in Futhark, and then used the comparison module to compare the
+ runs of the original model versus the Futhark implementation.
 
-% display results
+The comparison module displays the average execution time of each of the models,
+which lets me track exactly how much faster (or slower) the Futhark model has
+been running, compared to the original implementation.
 
-% discussion of futhark vs. python performance
-%% where does the performance gains come from?
+Furthermore, and ultimately most important, the comparison module also
+revealed the difference in results between the two models. 
+As I have precisely zero knowledge on the field of Small Angle Scatterings, 
+the only option for me regarding correctness of the models has been to take
+the results of the already existing models as the "correct" results of the
+computations. Consequently, any difference between my Futhark results and the
+original results has been direct indicators of faulty implementations or bugs
+somewhere along my work.
+Therefore, I have been developing and debugging my solution by repeatedly
+comparing it's results to that of the original implementation.
 
-% discussion of futhark vs. opencl performance
-%% where does the performance gains come from?
+Conversely, I have chosen to accept test runs with exactly no difference
+between the original and the Futhark implementation, as indicators of a
+correct implementation. 
+Test runs with thousands of data points defined as double-precision
+with more than ten decimal points have still resulted in a total difference of
+zero\footnote{see \ref{sec:correctness}} between the models.
+This is what I base my assertion of a correctness on, as even the smallest 
+mistakes along the calculations would surely have resulted in a very significant
+accumulated error in the final results.
+
+The test driven development has also lead to some of the biggest performance
+optimizations in my design of kernelfut.py.
+In example, the initial tests showed that the Futhark kernels were unreasonably
+much slower than the already existing kernels. After some investigation I found
+that this was because the original implementation buffered the input data in the
+OpenCL context long time before the actual kernel was called in Python, whereas
+the first versions of the Futhark kernel buffered the input data in the OpenCL
+context as part of the \texttt{run\_kernel} function call.
+
+\subsection{Futhark vs. Python performance}
+I have tested Futhark vs. Python performance on two different models; 
+the \texttt{line} model and the \texttt{broad\_peak} model.
+
+\subsubsection{\texttt{line} model}
+The \iq function for the line model is defined as $I(q) = A + B*q$, which is
+equivalent to one of the most basic high school math functions, $f(x) = ax + b$.
+
+\begin{figure}
+    \begin{tikzpicture}
+      \begin{axis}[
+        mbarplot,
+        title={1D performance},
+        xlabel={number of datapoints},
+        ylabel={Time in milliseconds},
+        width=0.9\textwidth,
+        height=3.5cm,
+        symbolic x coords={100,500,2500,5000,10000},
+        bar width=9pt,
+        enlargelimits=0.15,
+        ybar=10pt,% configures ‘bar shift’
+        bar width=12pt,
+        nodes near coords,
+        legend style={legend pos=north west}
+      ]
+      \addplot plot coordinates {(100, 0.24 ) (500, 0.24 ) (2500, 0.26 ) (5000, 0.26 ) (10000, 0.28 )};
+      \addplot plot coordinates {(100, 0.93 ) (500, 0.94 ) (2500, 0.97 ) (5000, 0.95 ) (10000, 0.97 )};
+      \legend{Current, Futhark}
+      \end{axis}
+    \end{tikzpicture}
+    \label{fig:line-graph}
+    \caption{Comparison between Python and OpenCL performance for simple model}
+\end{figure}
+
+In figure \ref{fig:line-graph} we see that for one dimensional calculations,
+both implementations keep a steady execution time. Futhark is consistently 
+slower, but that can be explained by the overhead of using the Futhark kernels.
+
+More interesting is it to look at the execution time of the 2D version of the 
+experiment. For all \sasmodels 2D experiments, we are not just calculating \iq
+for each q, but actually $\sum_{q \in Q} \sum_{q \in Q} Iq(q_x, q_y)$.
+So if we have $n$ q, we need to perform $n^2$ calculations
+
+n datapoints of q results in
+having to calculate the scattering for q squared .
+we have that we are actually 
+¨
+
+\subsubsection{Where does the performance gains come from?}
+
+\subsection{Futhark vs. OpenCL performance}
+\subsubsection{Where does the performance gains come from?}
 % show core_shell-whatever; and elaborate on loop flattening into maps
@@
 
-% correctness
+\subsection{Correctness of Futhark results}
 

### 2017-11-07 report/draft.tex
asd

@@
 \usepackage{pgfplots}
+\usepackage{hyperref}
 
@@
     \begin{tikzpicture}
+
       \begin{axis}[
@@
     \end{tikzpicture}
+    \begin{tikzpicture}
+      \begin{axis}[
+        mbarplot,
+        title={2D performance},
+        xlabel={number of datapoints},
+        ylabel={Time in milliseconds},
+        width=0.9\textwidth,
+        height=4cm,
+        symbolic x coords={100,500,2500,5000,10000},
+        bar width=9pt,
+        enlargelimits=0.15,
+        ybar=10pt,% configures ‘bar shift’
+        bar width=12pt,
+        nodes near coords,
+        legend style={legend pos=north west}
+      ]
+      \addplot plot coordinates {(100, 0.31 ) (500, 2.78 ) (2500, 127.01 ) (5000, 501.31 ) (10000, 1970.14 )};
+      \addplot plot coordinates {(100, 0.98 ) (500, 1.91 ) (2500, 22.73 ) (5000, 86.03 ) (10000, 338.10 )};
+
+      \legend{Current, Futhark}
+
+      \end{axis}
+    \end{tikzpicture}
     \label{fig:line-graph}
-    \caption{Comparison between Python and OpenCL performance for simple model}
+    \caption{Comparison between Python and Futhark performance for simple model}
 \end{figure}
@@
 for each q, but actually $\sum_{q \in Q} \sum_{q \in Q} Iq(q_x, q_y)$.
-So if we have $n$ q, we need to perform $n^2$ calculations
+So if we have $n$ q, we need to perform $n^2$ calculations.
 
-n datapoints of q results in
-having to calculate the scattering for q squared .
-we have that we are actually 
-¨
+We see in the figure, that Futhark at 10000^2 datapoints now greatly outperforms
+the Python model.
+
+\subsubsection{\texttt{broad\_peak} model}
+Broad Peak is quite a different beast than the line model from before. It's \iq
+function is defined as 
+$I(q) = \frac{A}{q^n} + \frac{C}{1 + (|q - q_0|\xi)^m} + B$.\footnote{see \url{http://www.sasview.org/docs/user/models/broad_peak.html}}
+
+\begin{figure}
+      \begin{tikzpicture}
+      \begin{axis}[
+        mbarplot,
+        title={1D performance},
+        xlabel={number of datapoints},
+        ylabel={Time in milliseconds},
+        width=0.9\textwidth,
+        height=3.5cm,
+        symbolic x coords={100,500,2500,5000,10000},
+        bar width=9pt,
+        enlargelimits=0.15,
+        ybar=10pt,% configures ‘bar shift’
+        bar width=12pt,
+        nodes near coords,
+        legend style={legend pos=north west}
+      ]
+      \addplot plot coordinates {(100,0.27) (500,0.30) (2500,0.45) (5000,0.63) (10000,1.22)};
+      \addplot plot coordinates {(100,1.22) (500,1.23) (2500,1.23) (5000,1.25) (10000,1.27)};
+
+      \legend{Current, Futhark}
+      \end{axis}
+    \end{tikzpicture}
+    \begin{tikzpicture}
+      \begin{axis}[
+        mbarplot,
+        title={2D performance},
+        xlabel={number of datapoints},
+        ylabel={Time in milliseconds},
+        width=0.9\textwidth,
+        height=4cm,
+        symbolic x coords={100,500,2500,5000,10000},
+        bar width=9pt,
+        enlargelimits=0.15,
+        ybar=10pt,% configures ‘bar shift’
+        bar width=12pt,
+        nodes near coords,
+        legend style={legend pos=north west}
+      ]
+      \addplot plot coordinates {(100,1.06 ) (500,21.50 ) (2500,675.12) (5000,2686.44 ) (10000,10687)};
+      \addplot plot coordinates {(100,1.27 ) (500,2.27 ) (2500,27.48) (5000,103.66 ) (10000,409)};
+
+      \legend{Current, Futhark}
+
+      \end{axis}
+    \end{tikzpicture}
+    \label{fig:broad-peak-graph}
+    \caption{Comparison between Python and Futhark performance for complex model}
+\end{figure}
+
+In figure \ref{fig:line-graph} we see that for one dimensional calculations,
+both implementations keep a steady execution time. Futhark is consistently 
+slower, but that can be explained by the overhead of using the Futhark kernels.
+
+More interesting is it to look at the execution time of the 2D version of the 
+experiment. For all \sasmodels 2D experiments, we are not just calculating \iq
+for each q, but actually $\sum_{q \in Q} \sum_{q \in Q} Iq(q_x, q_y)$.
+So if we have $n$ q, we need to perform $n^2$ calculations.
+
+We see in the figure, that Futhark at 10000^2 datapoints now greatly outperforms
+the Python model.
+\subsubsection{Where does the performance gains come from?}
+These comparisons are between Python/Numpy- and Futhark models.
+Although Numpy can execute both of the models as vectorized calculations, 
+it is still vastly outperformed by Futhark, which can run hundreds of threads 
+simultaniously on the GPU.
+
+## REFERENCER HER ##
+
+\subsection{Futhark vs. OpenCL performance}
+\subsubsection{The \texttt{DAB} model}
+The DAB model is relatively simple, as it is described as $I(q) = \frac{L^3}{1 + (qL)^2}$,
+requiring us to work with just one single variable.
+  \begin{figure}
+    \begin{tikzpicture}
+      \begin{axis}[
+        mbarplot,
+        title={1D performance},
+        xlabel={number of datapoints},
+        ylabel={Time in milliseconds},
+        width=0.9\textwidth,
+        height=3.5cm,
+        symbolic x coords={100,500,2500,5000,10000},
+        bar width=9pt,
+        enlargelimits=0.15,
+        ybar=10pt,% configures ‘bar shift’
+        bar width=12pt,
+        nodes near coords,
+        legend style={legend pos=north west}
+      ]
+      \addplot plot coordinates {(100, 0.54 ) (500, 0.74 ) (2500, 0.76 ) (5000, 0.74 ) (10000, 0.77 )};
+      \addplot plot coordinates {(100, 0.86 ) (500, 0.85 ) (2500, 0.87 ) (5000, 0.87 ) (10000, 0.90 )};
+
+      \legend{Current, Futhark}
+
+      \end{axis}
+    \end{tikzpicture}
+    \begin{tikzpicture}
+      \begin{axis}[
+        mbarplot,
+        title={2D performance},
+        xlabel={number of datapoints},
+        ylabel={Time in milliseconds},
+        width=0.9\textwidth,
+        height=4cm,
+        symbolic x coords={100,500,2500,5000,10000},
+        bar width=9pt,
+        enlargelimits=0.15,
+        ybar=10pt,% configures ‘bar shift’
+        bar width=12pt,
+        nodes near coords,
+        legend style={legend pos=north west}
+      ]
+      \addplot plot coordinates {(100, 0.78 ) (500, 1.61 ) (2500, 50.36 ) (5000, 196.19 ) (10000, 771.43 )};
+      \addplot plot coordinates {(100, 0.90 ) (500, 1.75 ) (2500, 26.25 ) (5000, 101.59 ) (10000, 402.69 )};
+
+      \legend{Current, Futhark}
+
+      \end{axis}
+    \end{tikzpicture}
+    \label{fig:dab-graph}
+    \caption{Comparison between OpenCL and Futhark performance for simple model}
+  \end{figure}
+
+\subsubsection{The \texttt{Core Shell Parallelepiped} model}
+The Core Shell Parallelepiped model is by far one of the most complex models
+in \sasmodels' catalogue.
+With it's more than ten parameters and it's two nested loops, it is 
+a candidate model for the Futhark compiler to attempt to optimize.
+  \begin{figure}
+    \begin{tikzpicture}
+      \begin{axis}[
+        mbarplot,
+        title={1D performance},
+        xlabel={number of datapoints},
+        ylabel={Time in milliseconds},
+        width=0.9\textwidth,
+        height=3.5cm,
+        symbolic x coords={100,500,2500,5000,10000},
+        bar width=9pt,
+        enlargelimits=0.15,
+        ybar=10pt,% configures ‘bar shift’
+        bar width=12pt,
+        nodes near coords,
+        legend style={legend pos=north west}
+      ]
+      \addplot plot coordinates {(100, 20.88 ) (500, 21.19 ) (2500, 21.99) (5000, 22.44 ) (10000, 42.99)};
+      \addplot plot coordinates {(100, 2.68 ) (500, 3.67 ) (2500, 8.00 ) (5000, 13.18 ) (10000,23.90 )};
+
+      \legend{Current, Futhark}
+
+      \end{axis}
+    \end{tikzpicture}
+    \begin{tikzpicture}
+      \begin{axis}[
+        mbarplot,
+        title={2D performance},
+        xlabel={number of datapoints},
+        ylabel={Time in milliseconds},
+        width=0.9\textwidth,
+        height=4cm,
+        symbolic x coords={100,500,2500,5000,10000},
+        bar width=9pt,
+        enlargelimits=0.15,
+        ybar=10pt,% configures ‘bar shift’
+        bar width=12pt,
+        nodes near coords,
+        legend style={legend pos=north west}
+      ]
+      \addplot plot coordinates {(100, 0.83 ) (500, 2.17 ) (2500, 61.66 ) (5000, 232.32 ) (10000, 986.73 )};
+      \addplot plot coordinates {(100, 2.04 ) (500, 2.88 ) (2500, 23.83 ) (5000, 91.87 ) (10000, 357.89 )};
+
+      \legend{Current, Futhark}
+
+      \end{axis}
+    \end{tikzpicture}
+    \label{fig:core-shell-graph}
+    \caption{Comparison between OpenCL and Futhark performance for complex model}
+  \end{figure}
 
 \subsubsection{Where does the performance gains come from?}
+While it was expected to see Futhark outperform the strictly CPU based 
+computations performed by Numpy in the Python models, it was more uncertain
+how much (or even whether) Futhark would outperform OpenCL.
 
-\subsection{Futhark vs. OpenCL performance}
-\subsubsection{Where does the performance gains come from?}
+However, data shows that Futhark does perform better. The increased performance 
+in figure \ref{fig:dab} is not really significant until we reach very high n,
+which can be attributed to the simplicity of the model. With such simple models,
+Futhark cannot really exploit any of the possible optimizations, but it can gain
+some advantage by doing FUTHARK MAGI, which actually ends up roughly halving the 
+execution time compared to the original implementation.
+
+However, when we work with the rewritten core shell parallelepiped, we see huge
+speedups even at n=100. This 
+FUTHARK-MAGI HOISTING OG TING PÅ MAPS
+
 % show core_shell-whatever; and elaborate on loop flattening into maps

### 2017-11-08 report/draft.tex
dd

@@
 \usepackage{hyperref}
+\usepackage[utf8]{inputenc}
+\usepackage{mdframed}
+\usepackage{float}
+\floatstyle{boxed}
+\restylefloat{figure}
+
+\lstset{
+  basicstyle=\tiny,
+  showspaces=false,
+  showstringspaces=false,
+  columns=flexible,
+  breaklines=true
+}
 
@@
 
+\newcommand{\sasmodels}{\textit{SasModels}}
+\newcommand{\sasview}{\textit{SasView}}
+\newcommand{\iq}{\textit{I(q)}}
+\newcommand{\foriq}{\texttt{for\_iq}}
+\newcommand{\hasiq}{\texttt{has\_iq}}
+\newcommand{\futhark}{\texttt{Futhark for SasModels}}
+\newcommand{\barheight}{5cm}
+
 \title{
   \vspace{3cm}
-  \Huge{Optimizing SasView models} \\
-  \Large{Enhancing performance by rewriting kernels from OpenCL/OpenMP to Futhark}
+  \Huge{\futhark}\\
+  \Large{Enhancing performance by rewriting kernels from Python/OpenCL to
+    Futhark}
 }
@@
 \author{
-  \Large{Mikkel Storgaard Knudsen}
-  \\ \texttt{mikkelstorgaard@gmail.com} \\
+  \Large{Mikkel Storgaard Knudsen}\\
+  \texttt{mikkelstorgaard@gmail.com}\\
   \\
-  \Large{Supervisor}
-  \\ Martin Elsman
-  \\ \texttt{mael@di.ku.dk} \\
+  \Large{Supervisor}\\
+  Martin Elsman\\
+  \texttt{mael@di.ku.dk}
 }
@@
 
-\newcommand{\sasmodels}{\textit{SasModels}}
-\newcommand{\sasview}{\textit{SasView}}
-\newcommand{\iq}{\textit{I(q)}}
-\newcommand{\foriq}{\texttt{for\_iq}}
-\newcommand{\hasiq}{\texttt{has\_iq}}
 \begin{document}
@@
 \subsection{Goal and motivation}
-\textit{SasView} is a Small Angle Scattering (SAS) analysis package for the analysis of
-1D and 2D scattering data directly in inverse space.
-As \textit{SasView} is given data, it performs its analyses by running the input data
-through theory models provided by the standalone module \textit{Sasmodels}.
-\\
-\\
-The \textit{Sasmodels} models utilizes a set of kernels, which are implemented
-in both OpenCL and OpenMP.
-However, benchmark
-comparisons\footnote{https://futhark-lang.org/performance.html} between the
-parallel algorithm library \textit{Thrust} (implemented in C++/CUDA), and the
-array programming language Futhark, shows that there might severe speed
-improvements to gain by running these model calculations in Futhark, instead
-of in OpenCL and OpenMP.
-\\
-\\
+\sasview is a Small Angle Scattering (SAS) analysis package for the
+analysis of 1D and 2D scattering data directly in inverse space.
+As \sasview is given data, it performs its analyses by running the
+input data through theory models provided by the standalone module
+\sasmodels.
+
+The \sasmodels models utilizes a set of kernels, which are implemented
+in both Python and OpenCL. However, benchmark
+comparisons\footnote{https://futhark-lang.org/performance.html}
+between the parallel algorithm library \textit{Thrust}
+(implemented in C++/CUDA), and the array programming language Futhark, shows
+that there might severe speed improvements to gain by running these model
+calculations in Futhark, instead of in OpenCL as they are now.
+
 The goal of this project is to explore the potential speed gains obtained by
 rewriting, and testing the resulting speed improvements of, a subset of the
-\textit{Sasmodels} models.
+\sasmodels models.
 The subset will be chosen so that there are at least
@@
 model in it.
-\\
-\\
+
 If rewriting the models results in a significant\footnote{I.e. a speed increase
   in order of at least a magnitude}, speedup, it will reasonable to believe that 
-  the rest of the models in \textit{Sasmodels} can be rewritten for a similar
+  the rest of the models in \sasmodels can be rewritten for a similar
   speedup.
@@
 by \sasmodels' users.
+måske ikke så meget mere her
 % mere
@@
 
-\section{Method and performance results}
+\section{Method}
+
 % description of built-in sasmodels comparison functionality
@@
 
-\subsection{Development method}
 The development of Futhark for \sasmodels has been largely test-driven.
@@
 
+\section{Performance results}
+\label{performance}
 \subsection{Futhark vs. Python performance}
@@
     \begin{tikzpicture}
-
       \begin{axis}[
@@
         width=0.9\textwidth,
-        height=3.5cm,
+        height=\barheight,
         symbolic x coords={100,500,2500,5000,10000},
@@
         width=0.9\textwidth,
-        height=4cm,
+        height=\barheight,
         symbolic x coords={100,500,2500,5000,10000},
@@
 
-We see in the figure, that Futhark at 10000^2 datapoints now greatly outperforms
+We see in the figure, that Futhark at $10000^2$ datapoints now greatly outperforms
 the Python model.
-\subsubsection{Where does the performance gains come from?}
+
+\begin{mdframed}[
+  frametitle={Why does Futhark perform faster than Python?},
+  nobreak=true
+  ]
 These comparisons are between Python/Numpy- and Futhark models.
@@
 ## REFERENCER HER ##
+\end{mdframed}
 
@@
 
-\subsubsection{Where does the performance gains come from?}
+  \begin{mdframed}[
+    frametitle={Why does Futhark perform faster than OpenCL?},
+    nobreak=true]
 While it was expected to see Futhark outperform the strictly CPU based 
@@
 
-% show core_shell-whatever; and elaborate on loop flattening into maps
+\# show core-shell-whatever; and elaborate on loop flattening into maps
 %%% MASSER AF REFERENCER TIL TROELS' PAPERS
-
-\subsection{Correctness of Futhark results}
+\end{mdframed}
 
 \section{Discussion, future work}
-% is this immediately useful? not on single fire runs with low nq
-% can be used for data streams / bundling
-% experience gains can lay groundwork for parallelising finding best match
+\label{sec:discussion}
+The performance tests in \ref{performance} unequivocally shows, that
+there are speedups to gain by running Futhark kernels in favor of Python or
+OpenCL kernels.
+In general, Futhark starts surpassing the performance of competing kernels, when
+sample sizes rise above n=INDSÆT N. However, these gains are currently
+irrelevant for the most \sasview use cases, as almost all
+experiments are carried out on sample sizes of between 100-500.
+However, meetings with external project owner Wojciech Potrzebowski lead to the
+suggestion that simple bundling of \sasview experiments would enable us
+to leverage the Futhark speed improvements directly.
 
-% user directed: ease of use,
-% (evt.) skønhedsfejl
+In example, one could imagine running an experiment that continuously sends
+sensor data for processing in \sasview / \sasmodels.
+Instead of running an experiment for each sample\footnote{of size n=100-500}
+from the stream, one could let \sasmodels bundle these input samples in packages
+of 10-50 samples
+instead, and calculate these collectively to leverage the speedups that are
+apparent on Futhark calculations on samples of larger sizes.
+
+Please note, that sample sizes might have an upper bound, depending on the
+memory available on the workstations running the experiments.
+
+\subsection{Finding best match in \sasview}
+\label{sec:best-match}
+In addition to bundling input data, Wojciech also did suggest that the Futhark
+speedups could be used to implement faster best-match-finding for \sasview
+experiments.
+
+Among other things, \sasview is used to test multiple models against a data set,
+to find out which of the models that actually fits the data most closely.
+Currently this is not done in an efficient manner.
+Therefore, it would be interesting to implement a way of running multiple models
+on the same data set in parallel.
+And given the current Futhark vs. OpenCL results, the implementation should
+pursued in Futhark, as it has already been shown to be optimal
+(compared to OpenCL).
+
+\subsection{Usability of Futhark models}
+It is relevant to discuss how useful Futhark models are to the \sasview user
+groups. For the users who merely wants to use the models, it is not necessary
+to understand the implementation itself at all, as the user facing part of
+the model is just the model info file, which is just the same as the model infos
+for Python and OpenCL models \ref{sec:using-futhark}.
+
+However, for the model developer, it is now also necessary to have at least a
+rudimentary understanding of functional programming.
+In example, developers will have to learn how to use maps in
+places where they would usually use loops.
+
+Furthermore, the current solution requires severe amounts of boilerplate code to
+be included in the futhark model file.
+This boilerplate is caused by the usage of Futhark's module system, and
+therefore not strictly part of defining i.e. the \iq function of a model.
+The boilerplate might become a distraction for the model developer, so
+there might be a smaller educational challenge to solve here as well.
 
 \section{Conclusion}
+To answer the questions raised in this project definition, it was first
+necessary to implement a \sasmodels kernel for Futhark models. This was done
+successfully and in a very non-destructive way in the already existing
+\sasmodels code base.
+
+The kernel itself (\texttt{kernelfut.py}) depends on the existing code base, but
+the total change of the original code is two lines: an extra case in the
+switch statement that decides which kernel type to run for a given model.
+
+For Futhark model performances increases themselves, the results in
+\ref{performances}, tells us that there are performance increase of between
+200- and 2000 percent, depending on the original model's language and
+complexity.
+
+These gains are only realised when the sample size for a given run is high
+enough, but section \ref{sec:discussion} has proposals for \sasmodels extensions
+that could help \sasmodels users to reach these sample sizes, by bundling
+experiments.
+
+\begin{mdframed}[
+  frametitle={Presentation at \sasview{} Code Camp},
+  nobreak=true
+  ]
+Together with my supervisor and Futhark designer Troels Henriksen,
+I was invited for an hour-long presentation and discussion of \futhark{} at the
+yearly \textit{SasView Code Camp}, held at the Copenhagen Bio Science Park.\\\\
+There was about 20 \sasview{} users and developers attending at the presentation,
+which was about the definition and implementation of \futhark{}, along with
+a presentation of the result performance boosts for selected models.\\\\
+The project was well-received by the audience, and there was an eager discussion
+of whether and how it could be possible to really exploit the Futhark
+optimizations.\\\\
+The primary suggestion was to leverage Futhark to find best-fitting models,
+as described in \ref{sec:best-match}, but bundling continuous input from
+a sensor to exploit Futhark speed ups at high n was also a feature that many
+of the attendants expressed an interest in.
+\end{mdframed}
+
 % nævn at projektet er blevet præsenteret for en faktisk brugergruppe,

### 2017-11-08 report/draft.tex
ia

@@
   columns=flexible,
-  breaklines=true
+  breaklines=true,
+  numbers=left
 }
@@
 \section*{Abstract}
+This report describes the design and implementation of a Futhark extension for
+the modelling engine \sasmodels.
+\sasmodels{} performs small angle scattering calculations for the \sasview{}
+modelling suite, by sending it's input to a range of calculation models
+implemented in Python and OpenCL.\\\\
+
+%%% whaaaat
 
@@
 
-\sasmodels has built-in default definitions for $I(q_x,q_y)$ if working
+\sasmodels{} has built-in default definitions for $I(q_x,q_y)$ if working
 with two-dimensional data, and other functions such as \texttt{form\_volume(q)}
@@
 of $I(q_x, q_y)$ or other default functions.
-An example of this can be seen in appendix dab or line.
-
-A minimum viable model is showed in figure \ref{fig:linemodel}, and a more
-complex model can be seen in appendix \ref{app:core_shell}
+An example of this can be seen in figure \ref{fig:linemodel}.
+A more complex model can be seen in appendix \ref{fig:core_shell}
 
@@
 
-%% figure
-
 A simple example of a Futhark model is the line module, 
@@
 \label{sec:line-fut}
+
 \begin{figure}
-  % hele line_futhark.fut
+  \lstinputlisting[]{../sasmodels/sasmodels/models/futharks/line_futhark.fut}
   \caption{The complete definition of the line model implemented in Futhark.}
-  \label{fig:line-fut}
+  \label{fig:linemodel-futhark-full}
 \end{figure}
-There a several things happening in this fig. \ref{fig:line-fut}, besides the 
+
+There a several things happening in this fig. \ref{fig:linemodel-futhark-full}, besides the 
 definition of the line Iq function.
@@
 \begin{figure}
-  % hele for_iq.fut
-  % forklar module type
-  % forklar module 
+  \lstinputlisting[]{../sasmodels/sasmodels/models/futharks/header/for_iq.fut}
   \caption{The Futhark template that is used by the individual models.}
@@
 
-\subsection*{DAB model}
 \begin{figure}
-  \lstinputlisting[language=Python]{../sasmodels/sasmodels/models/dab.py}
-  \label{fig:dab}
+  \lstinputlisting[language=Python, firstline=44,lastline=66]{../sasmodels/sasmodels/models/dab.py}
+  \label{fig:dabmodel}
   \caption{The DAB model}
@@
 
-\subsection*{broad\_peak model}
 \begin{figure}
-  \lstinputlisting[language=Python]{../sasmodels/sasmodels/models/broad_peak.py}
+  \lstinputlisting[language=Python, firstline=44, lastline=94]{../sasmodels/sasmodels/models/broad_peak.py}
+  \label{fig:broadpeakmodel}
   \caption{The broad_peak model}
@@
 \begin{figure}
-  \lstinputlisting[language=Python]{../sasmodels/sasmodels/models/line.py}
+  \lstinputlisting[language=Python, linerange={27-55,57-68}]{../sasmodels/sasmodels/models/line.py}
+  \label{fig:linemodel}
   \caption{The line model}
@@
 \begin{figure}
-  \lstinputlisting[]{../sasmodels/sasmodels/models/futharks/line_futhark.fut}
+  \lstinputlisting[firstline=2, lastline=18]{../sasmodels/sasmodels/models/futharks/line_futhark.fut}
   \label{fig:linemodel-futhark}
@@
 \begin{figure}
-  \lstinputlisting[language=Python]{../sasmodels/sasmodels/models/line_futhark.py}
+  \lstinputlisting[language=Python, firstline=27, lastline=49]{../sasmodels/sasmodels/models/line_futhark.py}
+  \label{fig:linemodelinfo-futhark}
   \caption{The line model info file designating a Futhark path}
@@
 
+\begin{figure}
+  \lstinputlisting[language=Python, linerange={131-162}]{../sasmodels/sasmodels/models/core_shell_parallelepiped.py}
+  \lstinputlisting[language=C, linerange={10-81}]{../sasmodels/sasmodels/models/core_shell_parallelepiped.c}
+  \label{fig:core_shell}
+  \caption{The original core\_shell\_parallelepiped model (Iq and form volume)}
+\end{figure}
+
 \end{document}

### 2017-11-08 report/draft.tex
kk

--- 2017-11-07 report/draft.tex
@@
 However, when we work with the rewritten core shell parallelepiped, we see huge
-speedups even at n=100. This 
-FUTHARK-MAGI HOISTING OG TING PÅ MAPS
-
-% show core_shell-whatever; and elaborate on loop flattening into maps
-%%% MASSER AF REFERENCER TIL TROELS' PAPERS
+speedups even at n=100. This speedup mainly comes from Futharks ## INDSÆT DET 
+DER FEDE NOGET MED MAPS.
+Looking at the definition of core_shell_parallelepiped_futhark, we see that 11
+parameters are extracted from the local\_values array. We could fear that this
+happens for every application of Iq, but the Futhark compiler actually performs
+HOISTING, which is an optimization that rewrites the program internally, so that
+these variable allocations are moved out of the Iq function, and are thus only
+performed once, instead of n times.
+# KIIIIIIIIIILDE
 
 \subsection{Correctness of Futhark results}
+There has been one small issue with the correctness of the Futhark models.
+As the Futhark models are generic by design, and instantiated with the 
+programmer's floating point type of choice, we can run the Futhark kernels
+with any precision we want.
+
+However, some of the original \sasmodels models have Iq functions that are 
+hardcoded to use double precision floats in places, and even sometimes mixes
+doubles and singles. 
+As these models plays fast and loose with the precision type, the Futhark models
+will inevitably end up showing some total difference between it's results and 
+the results from the built in functions.
+
+However, several \sasview users have assured me, that it is not really a 
+problem. The worst examples of these differences has been accumulated errors of
+between 10e-6 and 10e-7, which is well within the error range accepted by 
+the \sasview users.
 

### 2017-11-08 report/draft.tex

--- 2017-11-08 report/draft.tex
@@
 However, when we work with the rewritten core shell parallelepiped, we see huge
-speedups even at n=100. This 
-FUTHARK-MAGI HOISTING OG TING PÅ MAPS
+speedups even at n=100. This speedup mainly comes from Futharks ## INDSÆT DET 
+DER FEDE NOGET MED MAPS.
+Looking at the definition of core_shell_parallelepiped_futhark, we see that 11
+parameters are extracted from the local\_values array. We could fear that this
+happens for every application of Iq, but the Futhark compiler actually performs
+HOISTING, which is an optimization that rewrites the program internally, so that
+these variable allocations are moved out of the Iq function, and are thus only
+performed once, instead of n times.
+# KIIIIIIIIIILDE
 
-\# show core-shell-whatever; and elaborate on loop flattening into maps
-%%% MASSER AF REFERENCER TIL TROELS' PAPERS
-\end{mdframed}
+\subsection{Correctness of Futhark results}
+There has been one small issue with the correctness of the Futhark models.
+As the Futhark models are generic by design, and instantiated with the 
+programmer's floating point type of choice, we can run the Futhark kernels
+with any precision we want.
 
+However, some of the original \sasmodels models have Iq functions that are 
+hardcoded to use double precision floats in places, and even sometimes mixes
+doubles and singles. 
+As these models plays fast and loose with the precision type, the Futhark models
+will inevitably end up showing some total difference between it's results and 
+the results from the built in functions.
+
+However, several \sasview users have assured me, that it is not really a 
+problem. The worst examples of these differences has been accumulated errors of
+between 10e-6 and 10e-7, which is well within the error range accepted by 
+the \sasview users.
 \section{Discussion, future work}

### 2017-11-09 report/draft.tex
k

@@
     nobreak=true]
+
 While it was expected to see Futhark outperform the strictly CPU based 
@@
 # KIIIIIIIIIILDE
+\end{mdframed}
 

### 2017-11-09 report/draft.tex
upw

@@
 
-\section{Demonstration of use of finished project}
-As \sasmodels is rarely used on its own, but rather as the backbone of 
-\sasview's calculation backend, I am not going to write much about 
-\sasview usage.
-Instead, I will describe how Futhark models in \sasmodels are leveraged 
-by \sasmodels' users.
-måske ikke så meget mere her
-% mere
+\section{Demonstration}
+\begin{mdframed}[frametitle={Terminology:}]
+  The work of this project has been spread out across multiple parts
+  of the \sasmodels program. Here is a list of the distinct parts of \sasmodels
+  that has been worked with:
+  \begin{itemize}
+    \item \texttt{kernelfut}: The \sasmodels Futhark kernel module that wraps 
+    \sasmodels input data and sends it to be calculated in the actual Futhark 
+    kernel. \texttt{kernelfut} is an alternative to \texttt{kernelpy} and
+    \texttt{kernelcl} which are already available in \sasmodels.
+
+    \item Model info files: Every model in \sasmodels is defined in a Python 
+    model info file, which defines all the model's properties, like the model's
+    function itself, and a list of the parameters needed to run the model.
+    See figure \ref{fig:dabmodel}
+
+    \item Futhark models: A Futhark model is the model module that contains the
+    functions needed to run a given model. See the Futhark line model in fig. 
+    \ref{fig:fig:linemodel-futhark}
+
+    \item The Futhark kernel template, as shown in figure \ref{fig:for-iq}:
+    The Futhark kernel template contains the module that takes a Futhark
+    model as input, and returns a Futhark kernel that can be called by 
+    \texttt{kernelfut}.
+
+    \item Futhark kernels: The Python OpenCL library that is produced by 
+    compiling the Futhark kernel template with a Futhark model.
+
+    \item \sasmodels experiments: A \sasmodels experiment is the action of 
+    running \sasmodels with some input data and one or more models, to get 
+    the resulting output data. The output data is the selected models applied
+    to the input data.
+  \end{itemize}
+
+\end{mdframed}
 
 \subsection{Describing \sasmodels' models}
-A \sasmodels model info file (hereafter referred to as just a model info) is a 
-Python file, that defines at least two primary properties of a 
-calculation model \textit{m}:
 \begin{itemize}
-
   \item a function definition \texttt{I(q)}, which defines scattered intensity 
@@
 
-\begin{figure}
-  % fuld beskrivelse af ax+b-figur, med parametre a, b, og en Iqxy
-  \caption{model info file for modelling a line}\label{fig:linemodel}
-\end{figure}
-
 \subsection{Defining \iq in Python and OpenCL}
@@
 and Python model info files, except for in two properties.
-As show in figure \ref{fig:futhark_dab}, there has been added a flag 
-%CHECK THISSS!!!
-\texttt{model\_is\_futhark: true}, and \iq is now a path to a precompiled 
+As show in figure \ref{fig:linemodelinfo-futhark}, \iq is now a dictionary
+with the single flag \texttt{model\_is\_futhark} set to \texttt{True},
+and the property \texttt{futhark\_path} points to the location of the compiled
 Futhark kernel.
@@
 The user does not need to have any Futhark compiler installed if he merely wants
-to use the precompiled Futhark module: Futhark models for \sasmodels compiles to
-portable Python libraries which has numpy and pyopencl as their only 
-dependencies - and as \sasmodels has these dependencies as well, the user will
-already have these dependencies sorted out, if he has a working installation of 
-\sasview or \sasmodels.
+to use the Futhark kernel: \futhark compiles to portable Python libraries which 
+has numpy and pyopencl as their only dependencies - and as \sasmodels has these 
+dependencies as well, the user will already have these dependencies sorted out, 
+if he has a working installation of \sasview or \sasmodels.
 
@@
 
-Other models, such as core-shell %FIX\
-imports libraries so they are available in the Iq function.
-%% som vist i core shell, kan det også være nødvendigt at definere og importere 
-% ekstra biblioteker
+Other models, such as core-shell imports libraries so they are 
+available in the Iq function.
 
@@
 
-\section{Implementation of \texttt{kernelfut.py}}
-\sasmodels
+\section{Adding \texttt{kernelfut} to \sasmodels}
+The following process happens when running a \sasmodels{} experiment through the
+built-in comparison module:
+\begin{enumerate}
+  \item The experiment parameters\footnote{choice(s) of model, number of 
+  evaluations, sample size, number type and so on} for the experiment are used
+  to create a ParameterTable.
 
-% current architecture of Sasmodels engine
-% description of kernelcl kernels
+  \item The ParameterTable is used to load (or generate) a set of input data for
+   the kernel.
+
+  \item The ParameterTable is used to initialize a \sasmodels kernel using the
+  desired model. In case of OpenCL and Futhark kernels, the input data is 
+  buffered to the OpenCL context already at this point. This is done to leave
+  as much buffering out of the time measurements of \sasmodels as possible.
+
+  \item \sasmodels packs the model and the data in a DirectModel wrapper object.
+
+  \item The DirectModel is run the desired number of times in \sasmodels,
+  and finally returns the model output, the average running time, and the
+  output difference between the models that have been compared. 
+\end{enumerate}
+
+To design \texttt{kernelfut}, it was first necessary to ape the 
+\sasmodels calculation loop as a Futhark function. 
+I chose to replicate \texttt{kernelpy}'s \texttt{\_loops(...)} function in 
+Futhark, as it is fairly simple to rewrite the Python code in Futhark.
+\texttt{\_loops()} is shown in figure \ref{fig:kernelpy_loop}, and the
+comparable Futhark function in figure \ref{fig:kernelfut_loop}.
+
+Second, it was necessary to implement a bridge between \sasmodels and the 
+precompiled Futhark kernels. 
+When a \sasmodels experiment wants to use a Futhark kernel for it's calculation, it must dynamically import the kernel
+% gør når jeg kan koncentrere mig
+
 % (smaller) description of kernelcl kernels
@@
 
-% nævn at projektet er blevet præsenteret for en faktisk brugergruppe,
-% og fik god og konstruktiv feedback
-% det inspirerede til yderligere projekter/forbedringer indenfor sasmodels
-
 \section*{Appendices}
@@
 
+\begin{figure}
+  \lstinputlisting[language=Python, linerange={180-252}]{../sasmodels/sasmodels/kernelpy.py}
+  \label{fig:kernelpy_loop}
+  \caption{The Python kernel calculation loop for \sasmodels}
+\end{figure}
+
+\begin{figure}
+  \lstinputlisting[linerange={13-53}]{../sasmodels/sasmodels/models/futharks/header/for_iq.fut}
+  \label{fig:kernelfut_loop}
+  \caption{The Futhark kernel calculation loop, aping fig \ref{fig:kernelpy_loop}}
+\end{figure}
+
 \end{document}

### 2017-11-09 report/draft.tex
tyui

@@
 \usepackage{float}
+\usepackage{amsmath}
 \floatstyle{boxed}
@@
 \newcommand{\barheight}{5cm}
+\newcommand{\call}{\texttt{\_\_call\_\_}}
 
@@
 \subsection{Goal and motivation}
-\sasview is a Small Angle Scattering (SAS) analysis package for the
+\sasview{} is a Small Angle Scattering (SAS) analysis package for the
 analysis of 1D and 2D scattering data directly in inverse space.
-As \sasview is given data, it performs its analyses by running the
+As \sasview{} is given data, it performs its analyses by running the
 input data through theory models provided by the standalone module
-\sasmodels.
-
-The \sasmodels models utilizes a set of kernels, which are implemented
+\sasmodels{}.
+\\\\
+The \sasmodels{} models utilizes a set of kernels, which are implemented
 in both Python and OpenCL. However, benchmark
@@
 calculations in Futhark, instead of in OpenCL as they are now.
-
+\\\\
 The goal of this project is to explore the potential speed gains obtained by
 rewriting, and testing the resulting speed improvements of, a subset of the
-\sasmodels models.
+\sasmodels{} models.
 The subset will be chosen so that there are at least
@@
 model in it.
-
+\\\\
 If rewriting the models results in a significant\footnote{I.e. a speed increase
   in order of at least a magnitude}, speedup, it will reasonable to believe that 
-  the rest of the models in \sasmodels can be rewritten for a similar
+  the rest of the models in \sasmodels{} can be rewritten for a similar
   speedup.
@@
   The work of this project has been spread out across multiple parts
-  of the \sasmodels program. Here is a list of the distinct parts of \sasmodels
+  of the \sasmodels{} program. Here is a list of the distinct parts of \sasmodels
   that has been worked with:
   \begin{itemize}
-    \item \texttt{kernelfut}: The \sasmodels Futhark kernel module that wraps 
-    \sasmodels input data and sends it to be calculated in the actual Futhark 
+    \item \texttt{kernelfut}: The \sasmodels{} Futhark kernel module that wraps 
+    \sasmodels{} input data and sends it to be calculated in the actual Futhark 
     kernel. \texttt{kernelfut} is an alternative to \texttt{kernelpy} and
     \texttt{kernelcl} which are already available in \sasmodels.
+    The final kernel object is FutKernel.
 
-    \item Model info files: Every model in \sasmodels is defined in a Python 
+    \item Model info files: Every model in \sasmodels{} is defined in a Python 
     model info file, which defines all the model's properties, like the model's
@@
     functions needed to run a given model. See the Futhark line model in fig. 
-    \ref{fig:fig:linemodel-futhark}
+    \ref{fig:linemodel-futhark}
 
@@
 
-    \item \sasmodels experiments: A \sasmodels experiment is the action of 
-    running \sasmodels with some input data and one or more models, to get 
+    \item \sasmodels{} experiments: A \sasmodels{} experiment is the action of 
+    running \sasmodels{} with some input data and one or more models, to get 
     the resulting output data. The output data is the selected models applied
@@
 environment} string representation of the scattering function.
-
+\\\\
 As the scattering function \texttt{I(q)} and its parameters are the only 
@@
 has the option to leave the model info defined with just these six properties.
-
+\\\\
 \sasmodels{} has built-in default definitions for $I(q_x,q_y)$ if working
@@
 for working with voluminous scatterings.
-
+\\\\
 Therefore, it is optional for the developer to define a custom definition
@@
 
-\subsection{Defining \iq in Python and OpenCL}
-A \sasmodels \iq function can be written in either Python (which will be
+\subsection{Defining \iq{} in Python and OpenCL}
+A \sasmodels{} \iq{} function can be written in either Python (which will be
 executed in a Numpy context at runtime), or it can be written in C, which will
@@
 executed as a kernel in an OpenCL context, preferably on a GPU.
-
-A Python \iq can be seen in figure \ref{fig:linemodel}, whereas both DAB (figure
-\ref{fig:dab} and core\_shell(figure \ref{fig:core-shell}) is written in C.
+\\\\
+A Python \iq{} can be seen in figure \ref{fig:linemodel}, whereas both DAB 
+(figure
+\ref{fig:dabmodel} and core\_shell(figure \ref{fig:core_shell}) are written in C.
 Notice that the main difference between the models lies in the Iq property.
-For Python models, \iq is a callable Python function, whilst the C \iq is a
+For Python models, \iq{} is a callable Python function, whilst the C \iq{} is a
  string which is copy-pasted wholesale into aforementioned OpenCL template.
-
-But with the addition of Futhark models in \sasmodels adds a third way of
-using and defining a \sasmodels model.
+\\\\
+But with the addition of Futhark models in \sasmodels{} adds a third way of
+using and defining a \sasmodels{} model.
 
@@
 There is a difference between defining and using a Futhark model.
-The \sasview user who wishes to use a Futhark model in his or her work does 
+The \sasview{} user who wishes to use a Futhark model in his or her work does 
 not need more than the futhark model info file, and a copy of the compiled
@@
 and Python model info files, except for in two properties.
-As show in figure \ref{fig:linemodelinfo-futhark}, \iq is now a dictionary
+\\\\
+As shown in figure \ref{fig:linemodelinfo-futhark}, \iq{} is now a dictionary
 with the single flag \texttt{model\_is\_futhark} set to \texttt{True},
@@
 Futhark kernel.
-
+\\\\
 The user does not need to have any Futhark compiler installed if he merely wants
-to use the Futhark kernel: \futhark compiles to portable Python libraries which 
-has numpy and pyopencl as their only dependencies - and as \sasmodels has these 
+to use the Futhark kernel: \futhark{} compiles to portable Python libraries which 
+has numpy and pyopencl as their only dependencies - and as \sasmodels{} has these 
 dependencies as well, the user will already have these dependencies sorted out, 
-if he has a working installation of \sasview or \sasmodels.
+if he has a working installation of \sasview{} or \sasmodels{}.
 
@@
 It is the model programmers responsibility to fill in all three functions.
-
+\\\\
 A simple example of a Futhark model is the line module, 
@@
 Python/OpenCL:
-
+\\
 With both Python and OpenCL models, the programmer can leave out Iqxy and 
-form\_volume so that \sasmodels automatically fills in the default definition
+form\_volume so that \sasmodels{} automatically fills in the default definition
 for these two functions instead.
-
+\\\\
 The Futhark programmer does not have this leisure yet, and must define the two
@@
 functions away in a parametric library module.
-
+\\\\
 The less removable nuisance is that the programmer must explicitly extract the
@@
 in the order they appear in the model info parameter list.
-
 Iq functions in Python and OpenCL can be defined with any number of function 
 parameters, but in Futhark, we need to adhere to the type declaration of Iq 
-(see \ref{has-iq}), which means that Iq can take only two arguments; the 
-scattering vector \texttt{q}, and the list \texttt{local\_values} of parameters
-needed for the scattering.
-
+(see \texttt{has\_iq} definition in \ref{fig:for-iq}), which means that Iq can 
+take only two arguments; the scattering vector \texttt{q}, and the list 
+\texttt{local\_values} of parameters needed for the scattering.
+\\\\
 This makes for an ugly prelude to the actual body of the Iq function, as shown
 in fig\ref{fig:linemodel-futhark}, or more critically in models like 
-core-shell \ref{fig:core-shell}.
+core-shell \ref{fig:core_shell}.
 
@@
 does only need to define three functions to be able to compile a working 
-\sasmodels Futhark model.
+\sasmodels{} Futhark model.
 However, those three functions are merely used by a parametric Futhark module 
-\texttt{for\_iq}, which takes care of the actual execution of the 
+\foriq{}, which takes care of the actual execution of the 
 Futhark kernel.
@@
 
-There a several things happening in this fig. \ref{fig:linemodel-futhark-full}, besides the 
-definition of the line Iq function.
-
+There a several things happening in this fig. \ref{fig:linemodel-futhark-full}, 
+besides the definition of the line Iq function.
+\\
 \texttt{module line} takes either single- or double precision floats as a
@@
 instead of maybe having to define linef64- and linef32 modules.
-
+\\\\
 Then, two kernels (one for each of f32 and f64) are initialised by giving
@@
 as parameters, returning \texttt{kernel\_float32} and \texttt{kernel\_float64}.
-
+\\
 Given these two modules, we can finally define four entry points into the
-finished model. These are the four entry points that can be called by \sasmodels
-whenever a \sasmodels experiment needs to execute calculations in a futhark
-kernel.
+finished model. These are the four entry points that can be called by 
+\sasmodels{} whenever a \sasmodels{} experiment needs to execute calculations 
+in a Futhark kernel.
 Each of the entry points are simply passing on their call parameters to the 
@@
 is a necessary part of every Futhark model.
-
+\\\\
 Other models, such as core-shell imports libraries so they are 
@@
 \subsubsection{header/for\_iq.fut (the Futhark kernel template)}
-\texttt{for\_iq.fut} contains the module type declaration for the \hasiq module,
- and it contains the definition of the parametric \foriq module.
-
-The type declaration \hasiq ensures, that any model module that is
-used as a parameter for the \foriq module defines at least the functions
+\texttt{for\_iq.fut} contains the module type declaration for the \hasiq{} module,
+ and it contains the definition of the parametric \foriq{} module. 
+\\
+The type declaration \hasiq{} ensures, that any model module that is
+used as a parameter for the \foriq{} module defines at least the functions
 Iq, Iqxy and form\_volume, as these functions are necessary for the definition
 of the Futhark kernel.
-
-Next, we have the actual definition of the module \foriq:
-\foriq takes two parameters (a parametric Futhark model), and a real type 
+\\\\
+Next, we have the actual definition of the module \foriq{}:
+\\
+\foriq{} takes two parameters (a parametric Futhark model), and a real type 
 to instantiate it with.
@@
 \texttt{Iq} available throughout the rest of the kernel.
-
+\\\\
 Then finally, it uses the Iq module to define the two functions 
 \texttt{run\_kernel}, and \texttt{run\_kernel\_2d}, which defines the actual 
-\sasmodels computation procedure with the scattering functions injected by 
+\sasmodels{} computation procedure with the scattering functions injected by 
 the Iq module.
@@
 This happens at runtime in \sasmodels, which means that the kernels aren't
- actually compiled until right before \sasmodels needs them.
-
+ actually compiled until right before \sasmodels{} needs them.
+\\\\
 Futhark models eschews this design and opts instead to separate code generation
@@
 It would be simple to copy the design of the OpenCL kernels, and compile Futhark
-kernels on \sasmodels runtime, but this design was cast aside in favour of the
+kernels on \sasmodels{} runtime, but this design was cast aside in favour of the
 design described in sec \ref{sec:generating-futhark}.
-
+\\
 There are three reasons for that:
@@
   \item Removing code generation from runtime unequivocally decreases the time
-  it takes to run a \sasmodels calculation. Although time taken by code 
+  it takes to run a \sasmodels{} calculation. Although time taken by code 
   generation is not covered by \sasmodels' built-in benchmark modules, it should
@@
   \item Compiling a kernel just in time brings the risk of encountering 
-  Futhark compilation errors at \sasmodels runtime. 
+  Futhark compilation errors at \sasmodels{} runtime. 
   It is arguably a user experience nightmare to encounter compilation errors at 
@@
   on Numpy and Pyopencl} Python libraries means that Futhark models are 
-  completely portable, and as such can be used by any \sasmodels or \sasview 
+  completely portable, and as such can be used by any \sasmodels{} or \sasview{} 
   user, without them needing to have a Futhark compiler installed on their 
@@
 built-in comparison module:
+\begin{figure}
+
 \begin{enumerate}
@@
 
-  \item The ParameterTable is used to initialize a \sasmodels kernel using the
+  \item The ParameterTable is used to initialize a \sasmodels{} kernel using the
   desired model. In case of OpenCL and Futhark kernels, the input data is 
   buffered to the OpenCL context already at this point. This is done to leave
-  as much buffering out of the time measurements of \sasmodels as possible.
+  as much buffering out of the time measurements of \sasmodels{} as possible.
 
-  \item \sasmodels packs the model and the data in a DirectModel wrapper object.
+  \item \sasmodels{} packs the model and the data in a DirectModel wrapper object.
 
@@
 \end{enumerate}
+\end{figure}
 
 To design \texttt{kernelfut}, it was first necessary to ape the 
-\sasmodels calculation loop as a Futhark function. 
+\sasmodels{} calculation loop as a Futhark function. 
 I chose to replicate \texttt{kernelpy}'s \texttt{\_loops(...)} function in 
@@
 comparable Futhark function in figure \ref{fig:kernelfut_loop}.
+\\\\
+Second, it was necessary to implement a bridge between \sasmodels{} and the 
+precompiled Futhark kernels. 
+When \sasmodels{} wants to use a Futhark kernel for it's calculations, 
+it must dynamically import the kernel during the initialisation of the FutKernel
+object. Figure \ref{fig:retrieving-futhark-class} shows how FutKernel uses
+the path and model name from the model info file, to load and initialize the
+model class that is available in the compiled Futhark kernel.
+\\\\
+After the Futhark kernel has been initialised and made available to FutKernel,
+we can retrieve the kernel's OpenCL queue, and buffer our input data onto the 
+OpenCL device immediately, so we don't have to spend time doing this during 
+kernel execution itself.
+Doing this reduces the Futhark overhead during kernel execution by between
+200 and 800 milliseconds.
+\\\\
+Finally, the FutKernel's _form attribute is set. _form is the entry point from
+the Futhark kernel, as seen in figure \ref{fig:linemodel-futhark-full}.
 
-Second, it was necessary to implement a bridge between \sasmodels and the 
-precompiled Futhark kernels. 
-When a \sasmodels experiment wants to use a Futhark kernel for it's calculation, it must dynamically import the kernel
-% gør når jeg kan koncentrere mig
+%% what is form??
+\begin{figure}
+  \lstinputlisting[linerange={30-32, 55-56, 74-82, 121-126}, language=Python]{../sasmodels/sasmodels/kernelfut.py}
+  \label{fig:retrieving-futhark-class}
+  \caption{Injecting the precompiled Futhark kernel in the FutKernel object}
+\end{figure}
 
-% (smaller) description of kernelcl kernels
-% description of current kernelfut implementation
+\begin{figure}
+  \lstinputlisting[linerange={128-140}, language=Python]{../sasmodels/sasmodels/kernelfut.py}
+  \label{fig:buffering-kernelfut}
+  \caption{Buffering the input data to the Futhark kernel's queue}
+\end{figure}
+
+Like with \texttt{kernelcl} and \texttt{kernelpy}, FutKernel must also implement
+\call{} so that DirectModel can call the model and perform the
+calculations.
+This has been done by writing a simplified version of the OpenCL kernel's 
+\call{} function. 
+\\\\
+As seen in figure \ref{fig:kernelfut_call} and \ref{fig:kernelcl_call},
+the primary task of the call function is to pack an array of arguments, and
+pass them on as arguments for a kernel.
+As \call{} must return a numpy array, the get() function is called at the end
+of the Futhark kernel call, to convert the output OpenCL Array into a 
+numpy array.
+\\\\
+Note that kernelfut does not have the evaluation loop that kernelcl has.
+The loop still exists, but has been moved inside of the Futhark template 
+instead (see fig. \ref{fig:for-iq}).
 
 \section{Method}
-
-% description of built-in sasmodels comparison functionality
-\sasmodels contains a comparison module called \texttt{compare.py}, which
+\sasmodels{} contains a comparison module called \texttt{compare.py}, which
 lets the user run different calculations on the same dataset.
@@
 the module that has let me design Futhark models in the first place.
-
-The development of Futhark for \sasmodels has been largely test-driven.
-For any given already existing model in the \sasmodels catalogue that I have
+\\\\
+The development of Futhark for \sasmodels{} has been largely test-driven.
+For any given already existing model in the \sasmodels{} catalogue that I have
 wished to implement in Futhark, I have started out by defining an equivalent
@@
  runs of the original model versus the Futhark implementation.
-
+\\\\
 The comparison module displays the average execution time of each of the models,
@@
 been running, compared to the original implementation.
-
+\\\\
 Furthermore, and ultimately most important, the comparison module also
@@
 comparing it's results to that of the original implementation.
-
+\\\\
 Conversely, I have chosen to accept test runs with exactly no difference
@@
 zero\footnote{see \ref{sec:correctness}} between the models.
+\\
 This is what I base my assertion of a correctness on, as even the smallest 
@@
 accumulated error in the final results.
-
+\\\\
 The test driven development has also lead to some of the biggest performance
@@
 \section{Performance results}
-\label{performance}
+\label{sec:performance}
 \subsection{Futhark vs. Python performance}
@@
 slower, but that can be explained by the overhead of using the Futhark kernels.
-
+\\\\
 More interesting is it to look at the execution time of the 2D version of the 
-experiment. For all \sasmodels 2D experiments, we are not just calculating \iq
+experiment. For all \sasmodels{} 2D experiments, we are not just calculating \iq
 for each q, but actually $\sum_{q \in Q} \sum_{q \in Q} Iq(q_x, q_y)$.
 So if we have $n$ q, we need to perform $n^2$ calculations.
-
+\\\\
 We see in the figure, that Futhark at 10000^2 datapoints now greatly outperforms
@@
 slower, but that can be explained by the overhead of using the Futhark kernels.
-
+\\\\
 More interesting is it to look at the execution time of the 2D version of the 
-experiment. For all \sasmodels 2D experiments, we are not just calculating \iq
+experiment. For all \sasmodels{} 2D experiments, we are not just calculating \iq
 for each q, but actually $\sum_{q \in Q} \sum_{q \in Q} Iq(q_x, q_y)$.
 So if we have $n$ q, we need to perform $n^2$ calculations.
-
+\\\\
 We see in the figure, that Futhark at $10000^2$ datapoints now greatly outperforms
@@
 it is still vastly outperformed by Futhark, which can run hundreds of threads 
-simultaniously on the GPU.
-
-## REFERENCER HER ##
+simultaneously on the GPU. In comparison, Numpy runs on the CPU, which rarely 
+runs more than 10-20 threads simultaneously.
+\\
+For models like broad\_peak, Futhark gains a significant speedup by 
+hoisting\cite[sec 2.3]{fhpc17} the allocation of variables 
+(fig. \ref{fig:broadpeakmodel_futhark}) outside of
+the function call, so they are allocated as constants in the \foriq{} 
+\texttt{run\_kernel} function, right before mapping Iq.Iq over the q input.
+\\\\
+Hoisting is not uncommon for compilers to do at all, but it does make a 
+difference for \sasmodels{}.
+The compiler moves the number of variable allocations from
+$O(\text{length of q vector } \cdot \text{number of parameters})$ to 
+$O(\text{number of parameters})$.
 \end{mdframed}
@@
 how much (or even whether) Futhark would outperform OpenCL.
-
+\\\\
 However, data shows that Futhark does perform better. The increased performance 
-in figure \ref{fig:dab} is not really significant until we reach very high n,
-which can be attributed to the simplicity of the model. With such simple models,
-Futhark cannot really exploit any of the possible optimizations, but it can gain
-some advantage by doing FUTHARK MAGI, which actually ends up roughly halving the 
-execution time compared to the original implementation.
-
+in figure \ref{fig:dab-graph} is not really significant until we reach very high
+n, which can be attributed to the simplicity of the model.
+\\
+With such simple models, there are not many things that Futhark can do smarter 
+than the normal OpenCL compiler, besides the gains that are possible from
+the aforementioned hoisting.
+\\\\
 However, when we work with the rewritten core shell parallelepiped, we see huge
-speedups even at n=100. This speedup mainly comes from Futharks ## INDSÆT DET 
-DER FEDE NOGET MED MAPS.
-Looking at the definition of core_shell_parallelepiped_futhark, we see that 11
-parameters are extracted from the local\_values array. We could fear that this
-happens for every application of Iq, but the Futhark compiler actually performs
-HOISTING, which is an optimization that rewrites the program internally, so that
-these variable allocations are moved out of the Iq function, and are thus only
-performed once, instead of n times.
-# KIIIIIIIIIILDE
+speedups even at n=100. This speedup comes from the Futhark compiler's 
+nested \texttt{map} optimizations: the Iq function is mapped
+over the array of q values, and the Iq function body is mainly a nested 
+map-map-reduce-reduce, which makes the final nested function a 
+map-map-map-reduce-reduce. 
+\\
+As this construct can be flattened 
+and parallelised\cite[sec. 5]{pldi17}, the Futhark compiler ensures that the
+calculations within the maps over the Gauss arrays and q values are coalesced,
+we are guaranteed efficient memory access.
+For the original OpenCL implementation, we risk that different kernels are
+accessing memory locations that are physically far apart, so that the GPU
+has to handle the memory accesses individually, instead of being able to 
+combine them.
 \end{mdframed}
@@
 \subsection{Correctness of Futhark results}
+\label{sec:correctness}
 There has been one small issue with the correctness of the Futhark models.
@@
 with any precision we want.
-
-However, some of the original \sasmodels models have Iq functions that are 
+\\\\
+However, some of the original \sasmodels{} models have Iq functions that are 
 hardcoded to use double precision floats in places, and even sometimes mixes
@@
 the results from the built in functions.
-
-However, several \sasview users have assured me, that it is not really a 
+\\\\
+However, several \sasview{} users have assured me, that it is not really a 
 problem. The worst examples of these differences has been accumulated errors of
 between 10e-6 and 10e-7, which is well within the error range accepted by 
-the \sasview users.
+the \sasview{} users.
 \section{Discussion, future work}
@@
 sample sizes rise above n=INDSÆT N. However, these gains are currently
-irrelevant for the most \sasview use cases, as almost all
+irrelevant for the most \sasview{} use cases, as almost all
 experiments are carried out on sample sizes of between 100-500.
 However, meetings with external project owner Wojciech Potrzebowski lead to the
-suggestion that simple bundling of \sasview experiments would enable us
+suggestion that simple bundling of \sasview{} experiments would enable us
 to leverage the Futhark speed improvements directly.
-
+\\\\
 In example, one could imagine running an experiment that continuously sends
-sensor data for processing in \sasview / \sasmodels.
+sensor data for processing in \sasview{} / \sasmodels.
 Instead of running an experiment for each sample\footnote{of size n=100-500}
-from the stream, one could let \sasmodels bundle these input samples in packages
+from the stream, one could let \sasmodels{} bundle these input samples in packages
 of 10-50 samples
@@
 apparent on Futhark calculations on samples of larger sizes.
-
+\\\\
 Please note, that sample sizes might have an upper bound, depending on the
@@
 experiments.
-
-Among other things, \sasview is used to test multiple models against a data set,
+\\\\
+Among other things, \sasview{} is used to test multiple models against a data set,
 to find out which of the models that actually fits the data most closely.
@@
 \subsection{Usability of Futhark models}
-It is relevant to discuss how useful Futhark models are to the \sasview user
+It is relevant to discuss how useful Futhark models are to the \sasview{} user
 groups. For the users who merely wants to use the models, it is not necessary
@@
 for Python and OpenCL models \ref{sec:using-futhark}.
-
+\\\\
 However, for the model developer, it is now also necessary to have at least a
@@
 places where they would usually use loops.
-
+\\\\
 Furthermore, the current solution requires severe amounts of boilerplate code to
@@
 there might be a smaller educational challenge to solve here as well.
-
+\\\\
 \section{Conclusion}
 To answer the questions raised in this project definition, it was first
-necessary to implement a \sasmodels kernel for Futhark models. This was done
+necessary to implement a \sasmodels{} kernel for Futhark models. This was done
 successfully and in a very non-destructive way in the already existing
-\sasmodels code base.
-
+\sasmodels{} code base.
+\\\\
 The kernel itself (\texttt{kernelfut.py}) depends on the existing code base, but
@@
 switch statement that decides which kernel type to run for a given model.
-
+\\\\
 For Futhark model performances increases themselves, the results in
-\ref{performances}, tells us that there are performance increase of between
+sec \ref{performance}, tells us that there are performance increase of between
 200- and 2000 percent, depending on the original model's language and
 complexity.
-
+\\\\
 These gains are only realised when the sample size for a given run is high
-enough, but section \ref{sec:discussion} has proposals for \sasmodels extensions
-that could help \sasmodels users to reach these sample sizes, by bundling
+enough, but section \ref{sec:discussion} has proposals for \sasmodels{} extensions
+that could help \sasmodels{} users to reach these sample sizes, by bundling
 experiments.
-
+\\\\
 \begin{mdframed}[
@@
   \caption{The broad_peak model}
+\end{figure}
+
+\begin{figure}
+  \lstinputlisting[firstline=3, lastline=17]{../sasmodels/sasmodels/models/futharks/broad_peak_futhark.fut}
+  \label{fig:broadpeakmodel_futhark}
+  \caption{The broad_peak model in Futhark}
 \end{figure}
@@
 
+\begin{figure}
+  \lstinputlisting[linerange={541-588}, language=Python]{../sasmodels/sasmodels/kernelcl.py}
+  \label{fig:kernelcl_call}
+  \caption{Calling OpenCL kernel}
+\end{figure}
+
+\begin{figure}
+  \lstinputlisting[linerange={152-192}, language=Python]{../sasmodels/sasmodels/kernelfut.py}
+  \label{fig:kernelfut_call}
+  \caption{Calling FutKernel}
+\end{figure}
+
 \end{document}

### 2017-11-09 report/draft.tex
asd

@@
 \thispagestyle{empty}
-
+\clearpage
 \section*{Abstract}
-This report describes the design and implementation of a Futhark extension for
-the modelling engine \sasmodels.
 \sasmodels{} performs small angle scattering calculations for the \sasview{}
 modelling suite, by sending it's input to a range of calculation models
-implemented in Python and OpenCL.\\\\
-
-%%% whaaaat
+implemented in Python and OpenCL. 
+\\
+A Futhark module was written and added to the 
+existing \sasmodels{} codebase, to test whether Futhark could outperform the 
+already existing model languages used in \sasmodels{}. The Futhark models
+were performance tested against a range of the original models.
+All following results showed a speed increase of between 200 and 2000\%, 
+all without loss of precision.  Speedups were however only apparent at 
+unrealistically large input sizes.
+\\
+Future \sasmodels{} extensions are then suggested, so these speed increases can be 
+leveraged in daily \sasview{} usage.
 
@@
 \section{Introduction}
-\subsection{Goal and motivation}
 \sasview{} is a Small Angle Scattering (SAS) analysis package for the
 analysis of 1D and 2D scattering data directly in inverse space.
-As \sasview{} is given data, it performs its analyses by running the
+When \sasview{} is given input data, it performs its analyses by running the
 input data through theory models provided by the standalone module
@@
 in both Python and OpenCL. However, benchmark
-comparisons\footnote{https://futhark-lang.org/performance.html}
+comparisons\cite{futhark}
 between the parallel algorithm library \textit{Thrust}
@@
 \sasmodels{} models.
-The subset will be chosen so that there are at least
-one ``simple'' model, one model of intermediate complexity, and one complex
-model in it.
+The subset will be chosen so that there are for each of Python and OpenCL 
+one ``simple'' model, and a complex one.
 \\\\
 If rewriting the models results in a significant\footnote{I.e. a speed increase
-  in order of at least a magnitude}, speedup, it will reasonable to believe that 
+  in order of at least 100\%}, speedup, it will reasonable to believe that 
   the rest of the models in \sasmodels{} can be rewritten for a similar
@@
 \section{Demonstration}
-\begin{mdframed}[frametitle={Terminology:}]
+\begin{mdframed}[
+  frametitle={Terminology:},
+  nobreak=true
+  ]
   The work of this project has been spread out across multiple parts
-  of the \sasmodels{} program. Here is a list of the distinct parts of \sasmodels
-  that has been worked with:
+  of the \sasmodels{} program. Here is a list of the distinct parts of 
+  \sasmodels{} that has been worked with:
   \begin{itemize}
@@
     \texttt{kernelcl} which are already available in \sasmodels.
-    The final kernel object is FutKernel.
 
@@
 
-\subsection{Describing \sasmodels' models}
+All \sasmodels{} models are described in model info file. These model info 
+files has at least two properties:
 \begin{itemize}
@@
 of $I(q_x, q_y)$ or other default functions.
-An example of this can be seen in figure \ref{fig:linemodel}.
-A more complex model can be seen in appendix \ref{fig:core_shell}
+An example of this can be seen in fig. \ref{fig:linemodel}.
+A more complex model can be seen in fig. \ref{fig:core_shell}
 
@@
 For models like broad\_peak, Futhark gains a significant speedup by 
-hoisting\cite[sec 2.3]{fhpc17} the allocation of variables 
+hoisting\cite[sec 2.3]{pldi17} the allocation of variables 
 (fig. \ref{fig:broadpeakmodel_futhark}) outside of
@@
 \\
-As this construct can be flattened 
-and parallelised\cite[sec. 5]{pldi17}, the Futhark compiler ensures that the
-calculations within the maps over the Gauss arrays and q values are coalesced,
-we are guaranteed efficient memory access.
+As this construct can be flattened and parallelised\cite[sec. 5]{pldi17}, 
+we are can split our workload into many more tasks than the original OpenCL 
+implementation, which only has nq tasks. However, every single of these tasks
+consists of complex for-loop nests, and plenty of variable assignments.
+\\\\
+Furthermore, the Futhark compiler takes care to ensure that the calculations 
+over the Gauss arrays and q values are coalesced, so we are given 
+efficient memory access.
+\\
 For the original OpenCL implementation, we risk that different kernels are
@@
 \label{sec:discussion}
-The performance tests in \ref{performance} unequivocally shows, that
+The performance tests in \ref{sec:performance} unequivocally shows, that
 there are speedups to gain by running Futhark kernels in favor of Python or
@@
 For Futhark model performances increases themselves, the results in
-sec \ref{performance}, tells us that there are performance increase of between
+sec \ref{sec:performance}, tells us that there are performance increase of between
 200- and 2000 percent, depending on the original model's language and
@@
 \end{figure}
-
 \begin{figure}
@@
 
+\begin{thebibliography}{1}
+
+\bibitem{futhark}
+  T. Henriksen,
+  \textit{Futhark performances}
+  https://futhark-lang.org/performance.html
+
+\bibitem{pldi17}
+  T. Henriksen,
+  N. G. W. Serup,
+  M. Elsman,
+  F. Henglein,
+  C. E. Oancea,
+  \textit{Futhark: Purely Functional GPU-Programming with Nested Parallelism 
+  and In-Place Array Updates}
+  Department of Computer Science,
+  University of Copenhagen (Denmark),
+  2017.
+
+\end{thebibliography}
 \end{document}

### 2017-11-09 report/draft.tex
first done

@@
 \usepackage{amsmath}
+\usepackage{flafter}
 \floatstyle{boxed}
@@
 \newcommand{\sasview}{\textit{SasView}}
-\newcommand{\iq}{\textit{I(q)}}
+\newcommand{\iq}{\texttt{I(q)}}
 \newcommand{\foriq}{\texttt{for\_iq}}
@@
 \sasmodels{} performs small angle scattering calculations for the \sasview{}
-modelling suite, by sending it's input to a range of calculation models
+modeling suite, by sending its input to a range of calculation models
 implemented in Python and OpenCL. 
@@
 When \sasview{} is given input data, it performs its analyses by running the
-input data through theory models provided by the standalone module
+input data through theoretical models provided by the standalone module
 \sasmodels{}.
 \\\\
-The \sasmodels{} models utilizes a set of kernels, which are implemented
+\sasmodels{} utilizes a set of kernels, which are implemented
 in both Python and OpenCL. However, benchmark
-comparisons\cite{futhark}
+comparisons\cite[sec. 5.3]{array16}
 between the parallel algorithm library \textit{Thrust}
 (implemented in C++/CUDA), and the array programming language Futhark, shows
-that there might severe speed improvements to gain by running these model
-calculations in Futhark, instead of in OpenCL as they are now.
-\\\\
-The goal of this project is to explore the potential speed gains obtained by
+that there might be severe speed improvements to gain by running these model
+calculations in Futhark, instead of in OpenCL as they are now.\\
+\\\\
+The goal of \futhark{} is to explore the potential speed gains obtained by
 rewriting, and testing the resulting speed improvements of, a subset of the
-\sasmodels{} models.
+models in \sasmodels{}.
 The subset will be chosen so that there are for each of Python and OpenCL 
@@
 If rewriting the models results in a significant\footnote{I.e. a speed increase
-  in order of at least 100\%}, speedup, it will reasonable to believe that 
-  the rest of the models in \sasmodels{} can be rewritten for a similar
+  in order of at least 100\%} speedup, it will be reasonable to believe that 
+  the rest of the models in \sasmodels{} can be rewritten to obtain a similar
   speedup.
 
-\section{Demonstration}
+\section{Demonstration of the implementation}
 \begin{mdframed}[
@@
   ]
-  The work of this project has been spread out across multiple parts
+  \futhark{} is spread across multiple parts
   of the \sasmodels{} program. Here is a list of the distinct parts of 
@@
     \item Futhark models: A Futhark model is the model module that contains the
-    functions needed to run a given model. See the Futhark line model in fig. 
+    functions needed to run a given model. See the Futhark line model in figure 
     \ref{fig:linemodel-futhark}
@@
 \begin{itemize}
-  \item a function definition \texttt{I(q)}, which defines scattered intensity 
+  \item a function definition \iq{}, which defines scattered intensity 
   of the scattering vector \texttt{q}.
@@
   \item a list \texttt{parameters}, of the parameters that are used in 
-  \texttt{I(q)}. If \texttt{I(q)} does not take any parameters besides 
+  \iq{}. If \iq{} does not take any parameters besides 
   \texttt{q}, the list is left empty.
@@
 \\\\
-As the scattering function \texttt{I(q)} and its parameters are the only 
+As the scattering function \iq{} and its parameters are the only 
 two things that are strictly necessary in a model definition, the developer
@@
 \\\\
-\sasmodels{} has built-in default definitions for $I(q_x,q_y)$ if working
+\sasmodels{} has built-in default definitions for \texttt{I(qx,qy)} if working
 with two-dimensional data, and other functions such as \texttt{form\_volume(q)}
@@
 Therefore, it is optional for the developer to define a custom definition
-of $I(q_x, q_y)$ or other default functions.
-An example of this can be seen in fig. \ref{fig:linemodel}.
-A more complex model can be seen in fig. \ref{fig:core_shell}
+of \texttt{I(qx, qy)} or other default functions.
+An example of this can be seen in figure \ref{fig:linemodel}.
 
@@
 \\\\
-A Python \iq{} can be seen in figure \ref{fig:linemodel}, whereas both DAB 
-(figure
-\ref{fig:dabmodel} and core\_shell(figure \ref{fig:core_shell}) are written in C.
+A Python \iq{} can be seen in figure \ref{fig:linemodel}, whereas \iq{} for
+both DAB 
+(figure \ref{fig:dabmodel}) and core\_shell(figure \ref{fig:core_shell})
+are written in C.
 Notice that the main difference between the models lies in the Iq property.
 For Python models, \iq{} is a callable Python function, whilst the C \iq{} is a
- string which is copy-pasted wholesale into aforementioned OpenCL template.
-\\\\
-But with the addition of Futhark models in \sasmodels{} adds a third way of
-using and defining a \sasmodels{} model.
+ string which is copy-pasted wholesale into the aforementioned OpenCL template.
+\\
+With \futhark{}, a third way is introduced for using and defining a \sasmodels{} model.
 
@@
 The \sasview{} user who wishes to use a Futhark model in his or her work does 
-not need more than the futhark model info file, and a copy of the compiled
-Futhark kernel. In fact, the futhark model info file is identical to OpenCL- 
+not need more than the model info file, and the Futhark kernel.
+A Futhark model info file is identical to OpenCL- 
 and Python model info files, except for in two properties.
@@
 with the single flag \texttt{model\_is\_futhark} set to \texttt{True},
-and the property \texttt{futhark\_path} points to the location of the compiled
-Futhark kernel.
-\\\\
-The user does not need to have any Futhark compiler installed if he merely wants
+and the property \texttt{futhark\_path} points to the location of the Futhark kernel.
+\\\\
+The user does not need to have the Futhark compiler installed if he merely wants
 to use the Futhark kernel: \futhark{} compiles to portable Python libraries which 
-has numpy and pyopencl as their only dependencies - and as \sasmodels{} has these 
+has Numpy and Pyopencl as their only dependencies - and as \sasmodels{} has these 
 dependencies as well, the user will already have these dependencies sorted out, 
@@
 define it as a Futhark program and compile it using \texttt{futhark-pyopencl}.
-A model \textit{m} itself is defined as a parametric Futhark module\footnote{
+A Futhark model \textit{m} itself is defined as a parametric Futhark module\footnote{
   A more in-depth explanation of the complete definition of a Futhark module is 
-  in \ref{sec:generating-futhark}},
+  in sec \ref{sec:generating-futhark}},
 which contains an Iq-function, an Iqxy- and a form\_volume function.
@@
 \\\\
-A simple example of a Futhark model is the line module, 
-as depicted in figure \ref{fig:linemodel-futhark}.
-Note that Iqxy is non-default, whilst form\_volume is the default 
-implementation.
+A simple example is the line module, shown in figure \ref{fig:linemodel-futhark}.
+Note that Iqxy is a non-default definition, whilst form\_volume is the default 
+definition.
 
 \subsubsection{On the user friendliness of Futhark models}
-As shown in figure \ref{fig:linemodel-futhark}, there are currently two small 
+As apparent from figure \ref{fig:linemodel-futhark}, there are currently two small 
 nuisances in Futhark models, compared to equivalent models written in
@@
 \\\\
-The Futhark programmer does not have this leisure yet, and must define the two
-functions no matter what. This is admittedly a piece of annoying boiler plate,
-but it is also a thing that can be solved by compartmentalising the offending
-functions away in a parametric library module.
-\\\\
-The less removable nuisance is that the programmer must explicitly extract the
+The Futhark programmer does not have this leisure, but must therefore always
+define the two functions. This is boiler plate code, but in the future it could
+be solved by compartmentalising the offending functions away in a
+parametric library module.
+\\\\
+Less fixable is it that the programmer must explicitly extract the
 Iq function parameters from the list \texttt{local\_values}. This must be done
 in the order they appear in the model info parameter list.
+\\
 Iq functions in Python and OpenCL can be defined with any number of function 
-parameters, but in Futhark, we need to adhere to the type declaration of Iq 
-(see \texttt{has\_iq} definition in \ref{fig:for-iq}), which means that Iq can 
+parameters, but Futhark models must adhere to the type declaration of Iq 
+(see \texttt{has\_iq} definition in figure \ref{fig:for-iq}), which means that Iq can 
 take only two arguments; the scattering vector \texttt{q}, and the list 
@@
 \\\\
-This makes for an ugly prelude to the actual body of the Iq function, as shown
-in fig\ref{fig:linemodel-futhark}, or more critically in models like 
-core-shell \ref{fig:core_shell}.
+This leads to long preludes to the actual function body Iq function, as in
+figure \ref{fig:linemodel-futhark} (line model), or more glaringly like in
+figure \ref{fig:core_shell} (core shell model).
 
@@
 \label{sec:generating-futhark}
-As mentioned in \ref{sec:defining-futhark-model}, the futhark model developer
+As mentioned in sec. \ref{sec:defining-futhark-model}, the Futhark model developer
 does only need to define three functions to be able to compile a working 
@@
 
-There a several things happening in this fig. \ref{fig:linemodel-futhark-full}, 
-besides the definition of the line Iq function.
+Several things are happening in line\_futhark.fut besides (figure
+\ref{fig:linemodel-futhark-full}), besides the definition of the line Iq function.
+\\
 \\
@@
 Everything after the module line definition is boilerplate code, and could
-possibly defined in a less inelegant way in a later revision, but currently, it 
+possibly be defined in a more elegant way in a later revision, but currently, it 
 is a necessary part of every Futhark model.
 \\\\
-Other models, such as core-shell imports libraries so they are 
+Other models, such as core shell, imports libraries which are then
 available in the Iq function.
@@
 \subsubsection{header/for\_iq.fut (the Futhark kernel template)}
-\texttt{for\_iq.fut} contains the module type declaration for the \hasiq{} module,
+The file \texttt{for\_iq.fut} contains the module type declaration for the \hasiq{} module,
  and it contains the definition of the parametric \foriq{} module. 
@@
 \\\\
-Next, we have the actual definition of the module \foriq{}:
-\\
-\foriq{} takes two parameters (a parametric Futhark model), and a real type 
-to instantiate it with.
+Next, we have the actual definition of the module \foriq{}, which takes two parameters (a parametric Futhark model), and a real type to instantiate it with.
 It then instantiates the model with the real type, making the finished Iq module
 \texttt{Iq} available throughout the rest of the kernel.
-\\\\
+\\
 Then finally, it uses the Iq module to define the two functions 
@@
   \lstinputlisting[]{../sasmodels/sasmodels/models/futharks/header/for_iq.fut}
-  \caption{The Futhark template that is used by the individual models.}
+  \caption{The Futhark template that is used by the individual models. Notice
+    that this template is not copied, but appears only once in the system - it
+    is shared between the Futhark models.}
   \label{fig:for-iq}
@@
 It would be simple to copy the design of the OpenCL kernels, and compile Futhark
-kernels on \sasmodels{} runtime, but this design was cast aside in favour of the
+kernels at \sasmodels{} runtime, but this design was cast aside in favour of the
 design described in sec \ref{sec:generating-futhark}.
 \\
-There are three reasons for that:
+There are three reasons for this decision:
 \begin{enumerate}
@@
 \\\\
-Finally, the FutKernel's _form attribute is set. _form is the entry point from
+Finally, the FutKernel's \texttt{\_form} attribute is set. \texttt{\_form} is the entry point from
 the Futhark kernel, as seen in figure \ref{fig:linemodel-futhark-full}.
@@
   \lstinputlisting[linerange={30-32, 55-56, 74-82, 121-126}, language=Python]{../sasmodels/sasmodels/kernelfut.py}
+  \caption{Injecting the precompiled Futhark kernel in the FutKernel object}
   \label{fig:retrieving-futhark-class}
-  \caption{Injecting the precompiled Futhark kernel in the FutKernel object}
 \end{figure}
@@
   \lstinputlisting[linerange={128-140}, language=Python]{../sasmodels/sasmodels/kernelfut.py}
+  \caption{Buffering the input data to the Futhark kernel's queue}
   \label{fig:buffering-kernelfut}
-  \caption{Buffering the input data to the Futhark kernel's queue}
 \end{figure}
@@
 \\\\
-As seen in figure \ref{fig:kernelfut_call} and \ref{fig:kernelcl_call},
+As seen in figure \ref{fig:kernelfut_call} and figure \ref{fig:kernelcl_call},
 the primary task of the call function is to pack an array of arguments, and
@@
 The loop still exists, but has been moved inside of the Futhark template 
-instead (see fig. \ref{fig:for-iq}).
+instead (see figure \ref{fig:for-iq}).
 
@@
 revealed the difference in results between the two models. 
-As I have precisely zero knowledge on the field of Small Angle Scatterings, 
+As I have limited knowledge about the field of Small Angle Scatterings, 
 the only option for me regarding correctness of the models has been to take
@@
 original results has been direct indicators of faulty implementations or bugs
-somewhere along my work.
+somewhere in my work.
 Therefore, I have been developing and debugging my solution by repeatedly
@@
 with more than ten decimal points have still resulted in a total difference of
-zero\footnote{see \ref{sec:correctness}} between the models.
+zero\footnote{see sec. \ref{sec:correctness}} between the models.
 \\
@@
 optimizations in my design of kernelfut.py.
-In example, the initial tests showed that the Futhark kernels were unreasonably
-much slower than the already existing kernels. After some investigation I found
+In the example, the initial tests showed that the Futhark kernels were unreasonably
+slower than the already existing kernels. After some investigation I found
 that this was because the original implementation buffered the input data in the
@@
       \begin{axis}[
-        mbarplot,
         title={1D performance},
@@
       \begin{axis}[
-        mbarplot,
         title={2D performance},
@@
     \end{tikzpicture}
+    \caption{Comparison between Python and Futhark performance for simple model}
     \label{fig:line-graph}
-    \caption{Comparison between Python and Futhark performance for simple model}
 \end{figure}
@@
 both implementations keep a steady execution time. Futhark is consistently 
-slower, but that can be explained by the overhead of using the Futhark kernels.
+slower, but that can be explained by the overhead of launching the Futhark kernels.
 \\\\
@@
 for each q, but actually $\sum_{q \in Q} \sum_{q \in Q} Iq(q_x, q_y)$.
-So if we have $n$ q, we need to perform $n^2$ calculations.
-\\\\
-We see in the figure, that Futhark at 10000^2 datapoints now greatly outperforms
+So if we have a q vector of $n$ elements, we need to perform $n^2$ calculations.
+\\\\
+We see in the figure, that Futhark at $10000^2$ datapoints now greatly outperforms
 the Python model.
@@
 \subsubsection{\texttt{broad\_peak} model}
-Broad Peak is quite a different beast than the line model from before. It's \iq
+\texttt{broad\_peak} is quite different than the line model from before (see
+figure \ref{fig:broadpeakmodel}). Its \iq
 function is defined as 
@@
     \end{tikzpicture}
+    \caption{Comparison between Python and Futhark performance for complex model}
     \label{fig:broad-peak-graph}
-    \caption{Comparison between Python and Futhark performance for complex model}
 \end{figure}
@@
 both implementations keep a steady execution time. Futhark is consistently 
-slower, but that can be explained by the overhead of using the Futhark kernels.
+slower, but, again, that can be explained by the overhead of launching the Futhark kernels.
 \\\\
 More interesting is it to look at the execution time of the 2D version of the 
-experiment. For all \sasmodels{} 2D experiments, we are not just calculating \iq
+experiment. As explained earlier, all \sasmodels{} 2D experiments are not just calculating \iq
 for each q, but actually $\sum_{q \in Q} \sum_{q \in Q} Iq(q_x, q_y)$.
-So if we have $n$ q, we need to perform $n^2$ calculations.
+So if we have $n$ elements in $Q$ , we need to perform $n^2$ calculations.
 \\\\
@@
 hoisting\cite[sec 2.3]{pldi17} the allocation of variables 
-(fig. \ref{fig:broadpeakmodel_futhark}) outside of
+(figure \ref{fig:broadpeakmodel_futhark}) outside of
 the function call, so they are allocated as constants in the \foriq{} 
@@
 \subsubsection{The \texttt{DAB} model}
-The DAB model is relatively simple, as it is described as $I(q) = \frac{L^3}{1 + (qL)^2}$,
+The \texttt{DAB} model (figure \ref{fig:dabmodel}) is relatively simple, as it is described as $I(q) = \frac{L^3}{1 + (qL)^2}$,
 requiring us to work with just one single variable.
@@
     \end{tikzpicture}
+    \caption{Comparison between OpenCL and Futhark performance for simple model}
     \label{fig:dab-graph}
-    \caption{Comparison between OpenCL and Futhark performance for simple model}
   \end{figure}
@@
 \subsubsection{The \texttt{Core Shell Parallelepiped} model}
-The Core Shell Parallelepiped model is by far one of the most complex models
-in \sasmodels' catalogue.
-With it's more than ten parameters and it's two nested loops, it is 
+The \texttt{Core Shell Parallelepiped} model is by far one of the most complex models
+in \sasmodels' catalogue (figure \ref{fig:core_shell}). 
+With its more than ten parameters and its nested loops, it is 
 a candidate model for the Futhark compiler to attempt to optimize.
@@
     \end{tikzpicture}
+    \caption{Comparison between OpenCL and Futhark performance for complex model}
     \label{fig:core-shell-graph}
-    \caption{Comparison between OpenCL and Futhark performance for complex model}
   \end{figure}
@@
 \\\\
-However, some of the original \sasmodels{} models have Iq functions that are 
+However, some of the original model implementations \sasmodels{} have Iq functions that are 
 hardcoded to use double precision floats in places, and even sometimes mixes
@@
 \label{sec:discussion}
-The performance tests in \ref{sec:performance} unequivocally shows, that
+The performance tests in sec. \ref{sec:performance} unequivocally shows, that
 there are speedups to gain by running Futhark kernels in favor of Python or
@@
 \\\\
-In example, one could imagine running an experiment that continuously sends
+For example, one could imagine running an experiment that continuously sends
 sensor data for processing in \sasview{} / \sasmodels.
@@
 
-\subsection{Finding best match in \sasview}
+\subsection{Finding the best match in \sasview}
 \label{sec:best-match}
-In addition to bundling input data, Wojciech also did suggest that the Futhark
+In addition to bundling input data, Wojciech also suggested that the Futhark
 speedups could be used to implement faster best-match-finding for \sasview
@@
 And given the current Futhark vs. OpenCL results, the implementation should
-pursued in Futhark, as it has already been shown to be optimal
-(compared to OpenCL).
+be pursued in Futhark, as it has already been shown to be advantageous
+(compared to Open CL).
 
@@
 the model is just the model info file, which is just the same as the model infos
-for Python and OpenCL models \ref{sec:using-futhark}.
+for Python and OpenCL models (see sec. \ref{sec:using-futhark}).
 \\\\
@@
 rudimentary understanding of functional programming.
-In example, developers will have to learn how to use maps in
-places where they would usually use loops.
+In example, developers will have to learn how to use maps in Futhark in
+places where they would usually use loops in Python and C.
 \\\\
 Furthermore, the current solution requires severe amounts of boilerplate code to
-be included in the futhark model file.
+be included in the Futhark model file.
 This boilerplate is caused by the usage of Futhark's module system, and
-therefore not strictly part of defining i.e. the \iq function of a model.
-The boilerplate might become a distraction for the model developer, so
+therefore not strictly part of defining the \iq function of a model.
+For instance, the boilerplate might become a distraction for the model developer, so
 there might be a smaller educational challenge to solve here as well.
 \\\\
+\subsection{Why should \sasmodels{} developers use Futhark?}
+Currently, large scale data processing can be an unpleasant endeavor in
+\sasmodels{}, as some of the models (especially the Python based ones) have
+got runtimes that makes \sasview{} itself slow and irksome to work with.
+\\
+With Futhark models, the developers are assured that their models are being
+optimized by a compiler automatically - at least as long as they choose to
+follow sound functional programming patterns.\\
+All in all, \futhark{} could simply lead to a faster (and thereby nicer)
+experience for \sasview{} users.
+\\\\
+That Futhark models must be compiled before being available to \sasmodels{} at
+all, also aides the \sasmodels{} developer in designing his models. With the
+exception of out-of-bounds errors stemming from bad indexes into the parameters
+list, the Futhark type checker ensures that mistakes are discovered on compile time,
+instead of having to debug a C function through the stacktraces of by a faulty
+\sasmodels{} OpenCL kernel compilation.
+\\
+
 \section{Conclusion}
-To answer the questions raised in this project definition, it was first
+To answer the questions raised in the project definition, it has been 
 necessary to implement a \sasmodels{} kernel for Futhark models. This was done
@@
 The kernel itself (\texttt{kernelfut.py}) depends on the existing code base, but
-the total change of the original code is two lines: an extra case in the
+the total change of the original code is two lines, which is an extra case in the
 switch statement that decides which kernel type to run for a given model.
 \\\\
-For Futhark model performances increases themselves, the results in
-sec \ref{sec:performance}, tells us that there are performance increase of between
-200- and 2000 percent, depending on the original model's language and
-complexity.
-\\\\
+The \futhark{} tests in sec \ref{sec:performance} shows, that models
+in the \sasmodels{} catalogue can be ported to Futhark, with performance
+increases of between 200 and 2000 percent, depending on the original language
+(and the complexity) of the model.
+\\
 These gains are only realised when the sample size for a given run is high
-enough, but section \ref{sec:discussion} has proposals for \sasmodels{} extensions
+enough, but sec. \ref{sec:discussion} has proposals for \sasmodels{} extensions
 that could help \sasmodels{} users to reach these sample sizes, by bundling
@@
 \\\\
+I believe that \futhark{} should be elaborated upon, because its speedups can
+make it feasible to experiment with model fitting, streaming, and possibly other
+new features, that depends on low runtimes.
 \begin{mdframed}[
@@
 The primary suggestion was to leverage Futhark to find best-fitting models,
-as described in \ref{sec:best-match}, but bundling continuous input from
+as described in sec. \ref{sec:best-match}, but bundling continuous input from
 a sensor to exploit Futhark speed ups at high n was also a feature that many
@@
 
-\section*{Appendices}
-
-\begin{figure}
-  \lstinputlisting[language=Python, firstline=44,lastline=66]{../sasmodels/sasmodels/models/dab.py}
-  \label{fig:dabmodel}
-  \caption{The DAB model}
-\end{figure}
-
-\begin{figure}
-  \lstinputlisting[language=Python, firstline=44, lastline=94]{../sasmodels/sasmodels/models/broad_peak.py}
-  \label{fig:broadpeakmodel}
-  \caption{The broad_peak model}
-\end{figure}
-
-\begin{figure}
-  \lstinputlisting[firstline=3, lastline=17]{../sasmodels/sasmodels/models/futharks/broad_peak_futhark.fut}
-  \label{fig:broadpeakmodel_futhark}
-  \caption{The broad_peak model in Futhark}
-\end{figure}
-
-\begin{figure}
-  \lstinputlisting[language=Python, linerange={27-55,57-68}]{../sasmodels/sasmodels/models/line.py}
-  \label{fig:linemodel}
-  \caption{The line model}
-\end{figure}
-
-\begin{figure}
-  \lstinputlisting[firstline=2, lastline=18]{../sasmodels/sasmodels/models/futharks/line_futhark.fut}
-  \label{fig:linemodel-futhark}
-  \caption{The line model written in Futhark}
-\end{figure}
-
-\begin{figure}
-  \lstinputlisting[language=Python, firstline=27, lastline=49]{../sasmodels/sasmodels/models/line_futhark.py}
-  \label{fig:linemodelinfo-futhark}
-  \caption{The line model info file designating a Futhark path}
-\end{figure}
-
-\begin{figure}
-  \lstinputlisting[language=Python, linerange={131-162}]{../sasmodels/sasmodels/models/core_shell_parallelepiped.py}
-  \lstinputlisting[language=C, linerange={10-81}]{../sasmodels/sasmodels/models/core_shell_parallelepiped.c}
-  \label{fig:core_shell}
-  \caption{The original core\_shell\_parallelepiped model (Iq and form volume)}
-\end{figure}
-
-\begin{figure}
-  \lstinputlisting[language=Python, linerange={180-252}]{../sasmodels/sasmodels/kernelpy.py}
-  \label{fig:kernelpy_loop}
-  \caption{The Python kernel calculation loop for \sasmodels}
-\end{figure}
-
-\begin{figure}
-  \lstinputlisting[linerange={13-53}]{../sasmodels/sasmodels/models/futharks/header/for_iq.fut}
-  \label{fig:kernelfut_loop}
-  \caption{The Futhark kernel calculation loop, aping fig \ref{fig:kernelpy_loop}}
-\end{figure}
-\begin{figure}
-  \lstinputlisting[linerange={541-588}, language=Python]{../sasmodels/sasmodels/kernelcl.py}
-  \label{fig:kernelcl_call}
-  \caption{Calling OpenCL kernel}
-\end{figure}
-
-\begin{figure}
-  \lstinputlisting[linerange={152-192}, language=Python]{../sasmodels/sasmodels/kernelfut.py}
-  \label{fig:kernelfut_call}
-  \caption{Calling FutKernel}
-\end{figure}
-
 \begin{thebibliography}{1}
 
-\bibitem{futhark}
+\bibitem{array16}
   T. Henriksen,
-  \textit{Futhark performances}
-  https://futhark-lang.org/performance.html
+  C. E. Oancea,
+  K. F. Larsen,
+  \textit{Design and GPGPU Performance of Futhark's Redomap Construct},
+  2016
 
@@
 \end{thebibliography}
+
+\newpage
+\section*{Appendices}
+\begin{figure}
+  \lstinputlisting[language=Python, firstline=44,lastline=66]{../sasmodels/sasmodels/models/dab.py}
+  \caption{The DAB model}
+  \label{fig:dabmodel}
+\end{figure}
+
+\begin{figure}
+  \lstinputlisting[language=Python, firstline=44, lastline=94]{../sasmodels/sasmodels/models/broad_peak.py}
+  \caption{The broad\_peak model}
+  \label{fig:broadpeakmodel}
+\end{figure}
+
+\begin{figure}
+  \lstinputlisting[firstline=3, lastline=17]{../sasmodels/sasmodels/models/futharks/broad_peak_futhark.fut}
+  \caption{The broad\_peak model in Futhark}
+  \label{fig:broadpeakmodel_futhark}
+\end{figure}
+
+\begin{figure}
+  \lstinputlisting[language=Python, linerange={27-55,57-68}]{../sasmodels/sasmodels/models/line.py}
+  \caption{The line model}
+  \label{fig:linemodel}
+\end{figure}
+
+\begin{figure}
+  \lstinputlisting[firstline=2, lastline=18]{../sasmodels/sasmodels/models/futharks/line_futhark.fut}
+  \caption{The line model written in Futhark}
+  \label{fig:linemodel-futhark}
+\end{figure}
+
+\begin{figure}
+  \lstinputlisting[language=Python, firstline=27, lastline=49]{../sasmodels/sasmodels/models/line_futhark.py}
+  \caption{The line model info file designating a Futhark path}
+  \label{fig:linemodelinfo-futhark}
+\end{figure}
+
+\begin{figure}
+  \lstinputlisting[language=Python, linerange={131-162}]{../sasmodels/sasmodels/models/core_shell_parallelepiped.py}
+  \lstinputlisting[language=C, linerange={10-81}]{../sasmodels/sasmodels/models/core_shell_parallelepiped.c}
+  \caption{The original core\_shell\_parallelepiped model (Iq and form volume)}
+  \label{fig:core_shell}
+\end{figure}
+
+\begin{figure}
+  \lstinputlisting[language=Python, linerange={180-252}]{../sasmodels/sasmodels/kernelpy.py}
+  \caption{The Python kernel calculation loop for \sasmodels}
+  \label{fig:kernelpy_loop}
+\end{figure}
+
+\begin{figure}
+  \lstinputlisting[linerange={13-53}]{../sasmodels/sasmodels/models/futharks/header/for_iq.fut}
+  \label{fig:kernelfut_loop}
+  \caption{The Futhark kernel calculation loop, aping figure \ref{fig:kernelpy_loop}}
+\end{figure}
+\begin{figure}
+  \lstinputlisting[linerange={541-588}, language=Python]{../sasmodels/sasmodels/kernelcl.py}
+  \caption{Calling OpenCL kernel}
+  \label{fig:kernelcl_call}
+\end{figure}
+
+\begin{figure}
+  \lstinputlisting[linerange={152-192}, language=Python]{../sasmodels/sasmodels/kernelfut.py}
+  \caption{Calling FutKernel}
+  \label{fig:kernelfut_call}
+\end{figure}
 \end{document}
